---
title: "SBC for bayes factors (with rstan and bridgesampling)"
author: "Martin Modrák"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{SBC for bayes factors (with rstan and bridgesampling)}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

SBC can be used to verify Bayes factor computation. The key is to realize,
that whenever we compute a Bayes factor between two models $\mathcal{M}_0, \mathcal{M}_1$, with the associated prior $\pi^0_\text{prior}, \pi^1_\text{prior}$ and observational distributions $\pi^0_\text{obs}, \pi^1_\text{obs}$, it is the same as making an inference
over a larger model of the form:

$$
\begin{aligned}
i &\sim \text{Bernoulli}(\text{Pr}(\mathcal{M}_1))\\
\theta_i &\sim \pi^i_\text{prior} \\
y &\sim \pi_\text{obs}^i(\theta_i)
\end{aligned}
$$

The same model is also implied whenever we do Bayesian model averaging (BMA).
The Bayes factor is fully determined by the posterior distribution of the model index
$i$ in this larger model:

$$
\frac{\text{Pr}(\mathcal{M}_i \mid y)}{\text{Pr}(\mathcal{M}_j \mid y)} =
BF_{i,j}\frac{\text{Pr}(\mathcal{M}_i)}{\text{Pr}(\mathcal{M}_j)}
$$

This means, we can run SBC for this larger model! The results also naturally
generlize to multiple models.

Additionally, we may also employ diagnostics specific for binary variables,
like testing if the predictions are _calibrated_, e.g. that whenever posterior
probability of model 1 is 75%, we expect the model 1 to generate the data in about 75%
of the cases. We may also test the data-averaged posterior, i.e. that the average
posterior model probability is equal to the prior model probability.

## Setting up

Let's setup the environment (`bridgesampling` currently works only with `rstan`,
so we cannot use `cmdstanr` as do in other vignettes):

```{r setup, message=FALSE,warning=FALSE, results="hide"}
library(SBC)
library(ggplot2)
library(bridgesampling)
library(dplyr)
library(tidyr)

library(rstan)
rstan_options(auto_write = TRUE)

options(mc.cores = parallel::detectCores())

library(future)
plan(multisession)


cache_dir <- "./_bayes_factor_rstan_SBC_cache"

if(!dir.exists(cache_dir)) {
  dir.create(cache_dir)
}

fit_cache_dir <- file.path(cache_dir, "fits")
if(!dir.exists(fit_cache_dir)) {
  dir.create(fit_cache_dir)
}


theme_set(theme_minimal())
```


```{r enable progressr, eval = FALSE}
# Run this _in the console_ to report progress for all computations to report progress for all computations
# see https://progressr.futureverse.org/ for more options
progressr::handlers(global = TRUE)
```

We will use a model very similar to the `turtles` example from the 
`bridgesampling` package --- a test of presence of a random effect in
a binomial probit model. 
Specifically, under $H_1$

$$
\begin{aligned}
y_i &\sim  \text{Bernoulli}(\Phi(\alpha + \beta x_{i} + \gamma_{g_i})),  &i = 1,2,\ldots,N\\
\gamma_j &\sim N(0, \sigma),  &j = 1,2,\ldots, G \\
\alpha &\sim N(0, \sqrt{10}),\\
\beta &\sim N(0, \sqrt{10}),\\
\sigma &\sim \text{Half}N(0, 1)
\end{aligned}
$$

and under $H_0$ we assume that there is no random effect, i.e. that $\tau = 0$.

## Incorrect model and BF


Here is a naive implementation of the model under $H_0$:

```{r, comment = ""}
cat(readLines("stan/turtles_H0.stan"), sep = "\n")
```

and under H1:

```{r, comment = ""}
cat(readLines("stan/turtles_H1_bad_normalization.stan"), sep = "\n")
```

as the name of the file suggests, the implementation is in fact wrong. 
We will let you guess what's wrong and see if we can uncover this with SBC and
discuss the problem later.



```{r compile models}
m_H0 <- stan_model("stan/turtles_H0.stan")
m_H1_bad <- stan_model("stan/turtles_H1_bad_normalization.stan")
```

The simplest we can do is to actually write simulator that can simulate
the data and parameters under each hypothesis separately, instead of
directly simulating the BMA supermodel as it will let the `SBC` package
do some useful stuff for us.

We will also reject low variability datasets - see `vignette("rejection_sampling")`
for background. To gain maximum sensitivity, our simulator also
implements the log-likelihood (`lp__`) as a derived test quantity, including
all Jacobian adjustments.

```{r}
data("turtles", package = "bridgesampling")

sim_turtles <- function(model, N, C) {
  stopifnot(N >= 3 * C)

  ## Rejection sampling to avoid low-variability datasets
  for(i in 1:200) {

    alpha0_raw <- rnorm(1) 
    alpha1_raw <- rnorm(1) 
    alpha0 <- sqrt(10) * alpha0_raw
    alpha1 <- sqrt(10) * alpha1_raw
    
    clutch <- rep(1:C, length.out = N)
    x <- rnorm(N) / 3  
    
    log_lik_shared <- dnorm(alpha0_raw, log = TRUE) +
      dnorm(alpha1_raw, log = TRUE)
    
    if(model == 0) {
      predictor <- alpha0 + alpha1 * x
      log_lik_spec <- 0
    } else {
      sigma_clutch <- abs(rnorm(1))
      
      b_clutch_raw <- rnorm(C)
      b_clutch <- b_clutch_raw * sigma_clutch
      
      predictor <- alpha0 + alpha1 * x + b_clutch[clutch]
      
      log_lik_spec <- 
        log(sigma_clutch) + dnorm(sigma_clutch, log = TRUE) + log(2) +
        sum(dnorm(b_clutch_raw, log = TRUE))
    }
    
    prob <- pnorm(predictor)
    y <- rbinom(N, p = prob, size = 1)
    if(mean(y == 0) < 0.9 && mean(y == 1) < 0.9) {
      break
    }
  }
  if(i >= 200) {
    warning("Could not generate nice dataset")
  }
  
  log_lik_predictor <- sum(dbinom(y, size = 1, p = prob, log = TRUE))
  
  variables <- list(
      alpha0 = alpha0,
      alpha1 = alpha1,
      lp__ = log_lik_shared + log_lik_spec + log_lik_predictor
    )
  
  if(model == 1) {
    variables$sigma <- sigma_clutch
    variables$b <- b_clutch
  } 

  list(
    generated = list(
      N = N,
      y = y,
      x = x,
      C = C,
      clutch = clutch
    ),
    variables = variables
  )
}
```

We now generate some datasets for each model separately and then use 
`SBC_datasets_for_bf` to build a combined dataset sampling the true model
randomly, but that also keeps track of all the shared (and not shared) parameters:

```{r}
set.seed(54223248)
N_sims <- 200
N <- 72
C <- 8
ds_turtles_m0 <- generate_datasets(SBC_generator_function(sim_turtles, model = 0, N = N, C = C), n_sims = N_sims)
ds_turtles_m1 <- generate_datasets(SBC_generator_function(sim_turtles, model = 1, N = N, C = C), n_sims = N_sims)
ds_turtles <- SBC_datasets_for_bf(ds_turtles_m0, ds_turtles_m1)
```

Let us now build a `bridgesampling` backend --- this will take backends
based on the individual models as arguments. The backend in question
need to implement the `SBC_fit_to_bridge_sampler()` S3 method. Out of the
box, this is supported by `SBC_backend_rstan_sample()` and `SBC_backend_brms()`.
Note that we keep
a pretty high number of iterations as `bridgesampling` needs that (the specific
numbers are taken straight from `bridgesampling` documentation).

One more trick we showcase is that we can cache individual model fits, if
we wrap a backend with `SBC_backend_cached()`. Since we will reuse the same
H0 model later, we will do this here just for the H0 model. Obviously everything
will also work without caching.

```{r}
iter <- 15500
warmup <- 500
init <- 0
backend_turtles_bad <- SBC_backend_bridgesampling(
  SBC_backend_cached(
    fit_cache_dir,
    SBC_backend_rstan_sample(m_H0, iter = iter, warmup = warmup, init = init)
    ),
  SBC_backend_rstan_sample(m_H1_bad, iter = iter, warmup = warmup, init = init)
)
res_turtles_bad <- compute_SBC(ds_turtles, backend_turtles_bad, 
                           keep_fits = FALSE,
                           cache_mode = "results",
                           cache_location = file.path(cache_dir, "turtles_bad.rds"))
```

Lets look at the default SBC diagnostics. Note that the plot by default
combines all parameters that are present in at least a single model. 
It then uses the implied BMA supermodel to get draws of the parameters (when
a parameter is not present in the model, all its draws are assumed to be `-Inf`).
This means that any SBC diagnostics for model parameters combines the 
correctness of the model index (and hence Bayes factor) with the
correctness of the individual model posteriors.

For clarity, we combine all the random effects in a 
single panel:

```{r ecdf-bad}
plot_ecdf_diff(res_turtles_bad, combine_variables = combine_array_elements)
```

OK, this is not great, but not completely conclusive --- given the number
of variables, some ECDFs being borderline outside the expected range is possible
even when the model works as expected. If we had no better tools, it would make 
sense to run more simulations at this point. But we do have better tools.

How about binary prediction calibration? Let's look at 
calibration plot (the default type relies on `reliabilitydiag::reliabilitydiag()`
and especially with smaller number of simulations, using `region.method = "resampling"`
is more reliable than the default approximation for the uncertainty region).
The histogram at the bottom shows the distribution of observed posterior
model probabilities.

```{r calib-bad}
plot_binary_calibration(res_turtles_bad, region.method = "resampling")
```

That is very clearly bad (the observed calibration is the black line, while the blue area is a 
confidence region assuming the probabilities are calibrated, we get miscalibration in multiple regions). We may also run a formal miscalibration test --- we first extract
the binary probabilities and then pass them to the test:

```{r}
bp_bad <- binary_probabilities_from_stats(res_turtles_bad$stats)
miscalibration_resampling_test(bp_bad$prob, bp_bad$simulated_value)
```


Even more bad. A lesson here is that the formal test is often a bit more sensitive than the test implied by the calibration plot --- in effect, the plot will show specific
model probabilities where the predictions are miscalibrated, while the test looks
at miscalibration as a global phenomenon, so it can integrate weak signals at 
different model probabilities. We also implement miscalibration test based
on Brier score, which is typically somewhat less sensitive, but included
here for completeness:

```{r}
brier_resampling_test(bp_bad$prob, bp_bad$simulated_value)
```


Finally, we may also look at the data-averaged posterior. If we believe we made
enough simulations for the 
central limit theorem to kick in, the highest sensitivity is
achieved by one sample t-test (we simulated with prior model probability = 0.5):

```{r}
t.test(bp_bad$prob, mu = 0.5)
```

Also very bad. In the cases where we are not sure the CLT holds (small number of simulations + "ugly" distribution of posterior model probabilities), we may also run the Gaffke test,
which makes no assumptions about the distribution of posterior model probabilities
beyond them being bounded between 0 and 1:

```{r}
gaffke_test(bp_bad$prob, mu = 0.5)
```

the Gaffke test has somewhat lower power (since it makes fewer assumptions), but the results are also very bad.


At this point, it would be tempting to blame the `bridgesampling` package,
but we cannot be sure. It is possible, that the problem is that one of the
models produces incorrect posterior.

An advantage of
using `SBC_datasets_for_bf` to generate the datasets, is that we can use the
same simulations to also do SBC for individual models (this and other
special handling is achieved internally via setting `var_attributes()`):

```{r ecdf-bad-split}
stats_split_bad <- split_SBC_results_for_bf(res_turtles_bad)
plot_ecdf_diff(stats_split_bad$stats_H0) + ggtitle("M0")
plot_ecdf_diff(stats_split_bad$stats_H1, 
               combine_variables = combine_array_elements) + ggtitle("M1")  
```

Those don't look bad (though we don't have a ton of simulation for each to be
sure).

We'll now reveal the problem (if you haven't already noticed). For normal
posterior sampling, Stan only needs the model to implement the density
up to a constant factor. However, for `bridgesampling`, we need all of those
constants included to obtain a correct estimate of the marginal likelihood.
In the H1 model, when we have `target += normal_lpdf(sigma | 0, 1);` we
are ignoring a factor of two as `sigma` has a half-normal distribution.

In situations like this, when the Bayes factor computation is 
biased, data-averaged posterior tends to be the most sensitive
check. However, for other problems in Bayes factor computation, like
incorrect variance or more subtle issues, testing binary prediction calibration
tends to be the most sensitive. Default SBC is less sensitive in the sense
that it typically requires more simulations to signal a problem than
binary prediction calibration checks, but more 
sensitive in the sense that some problems won't manifest with data-averaged posterior 
or binary prediction calibration, but will (with enough simulation and 
good test quantities) be detected by SBC.

[Tsukamura and Okada (2024)](https://doi.org/10.1007/s00407-022-00298-3) have noted
that many published Stan models do not include proper normalization for bounded
parameters and all of those models could thus produce incorrect Bayes factors
if used with `bridgesampling`, so this is a compuational problem
one can expect to actually see in the wild.

## Correct model

With the normalization constant included the correct H1 model is:

```{r, comment = ""}
cat(readLines("stan/turtles_H1.stan"), sep = "\n")
```

```{r}
m_H1 <- stan_model("stan/turtles_H1.stan")
```

Now lets run SBC with the correct model:

```{r}
backend_turtles <- SBC_backend_bridgesampling(
  SBC_backend_cached(
    fit_cache_dir,
    SBC_backend_rstan_sample(m_H0, iter = iter, warmup = warmup, init = init)
    ),
  SBC_backend_rstan_sample(m_H1, iter = iter, warmup = warmup, init = init)
)
res_turtles <- compute_SBC(ds_turtles, backend_turtles, 
                           keep_fits = FALSE,
                           cache_mode = "results",
                           cache_location = file.path(cache_dir, "turtles.rds"))
```



```{r ecdf-ok}
plot_ecdf_diff(res_turtles, combine_variables = combine_array_elements)
```

Wee see no big problems (although `sigma` is borderline, we would expect
this to disappear with more simulations, as we are quite sure the model
is in fact correct and `bridgesampling` works for this problem)

Similarly, binary prediction calibration also looks almost good

```{r calib-ok}
plot_binary_calibration(res_turtles, region.method = "resampling")
```

Formal miscalibration test shows no problems:

```{r}
bp <- binary_probabilities_from_stats(res_turtles$stats)
miscalibration_resampling_test(bp$prob, bp$simulated_value)
brier_resampling_test(bp$prob, bp$simulated_value)
```


And that data-averaged posterior? Both with t-test and the Gaffke test we see
no problems and resonably tight confidence intervals centered at the expected
value of 0.5.

```{r}
t.test(bp$prob, mu = 0.5)
gaffke_test(bp$prob, mu = 0.5)
```

So no problems here, up to the resolution available with `r N_sims` simulations.
In practice, we would recommend running more simulations for thorough checks,
but this is just an example that should compute in reasonable time.
