---
title: "SBC for brms models"
author: "Martin Modrák"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{SBC for brms models}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, message=FALSE,warning=FALSE, results="hide"}
library(SBC)
library(brms)
library(ggplot2)
options(brms.backend = "cmdstanr")
# options(brms.backend = "rstan") # Uncomment to use rstan instead

# Using parallel processing
library(future)
plan(multisession)

# The fits are very fast,
# so we force a minimum chunk size to reduce overhead of
# paralellization and decrease computation time.
options(SBC.min_chunk_size = 5)

# Setup caching of results
cache_dir <- "./brms_SBC_cache"
if(!dir.exists(cache_dir)) {
  dir.create(cache_dir)
}

```

Build a generator using brms and its `sample_prior = "only"` feature.

```{r}
# We need a "template dataset" to let brms build the model.
# The predictor (x) values will be used for data generation,
# the response (y) values will be ignored, but need to be present and 
# of the correct data type
template_data = data.frame(y = rep(0, 15), x = rnorm(15))
priors <- prior(normal(0,1), class = "b") +
  prior(normal(0,1), class = "Intercept") +
  prior(normal(0,1), class = "sigma")
generator <- SBC_generator_brms(y ~ x, data = template_data, prior = priors, 
                                thin = 50, warmup = 10000, refresh = 2000,
                                # Will generate the log density - this is useful, 
                                #but a bit computationally expensive
                                generate_lp = TRUE 
                                )
```



```{r}
set.seed(22133548)
datasets <- generate_datasets(generator, 100)
```

```{r}
# Reuse the compiled model and other info from the generator
backend <- SBC_backend_brms_from_generator(generator, chains = 1, thin = 1,
                            init = 0.1)

# More verbose alternative that results in exactly the same backend:
# backend <- SBC_backend_brms(y ~ x, template_dataset = template_data, prior = priors, warmup = 500, iter = 1000, chains = 1, thin = 1
#                            init = 0.1)
```


Compute the actual results

```{r}
results <- compute_results(datasets, backend, thin_ranks = 10, 
                    cache_mode = "results", 
                    cache_location = file.path(cache_dir, "first"))
```

And show ecdf_diff plot - there are no big problems at this resolution.

```{r results_plots}
plot_rank_hist(results)
plot_ecdf_diff(results)
```

## Using custom simulation code


Let's take a bit more complex model - with a single varying intercept.

This time we will not use the `brms` model to also simulate from prior, but 
simulate using an R function.

This  allows us to have different covariate values for each dataset, potentially improving sensitivity.

Let's take a Gaussian model with a single varying intercept.


The data can be generated using the following code:

```{r}
one_dataset_generator <- function(N, K) {
  # N - number of datapoints, K number of groups for the varying intercept
  stopifnot(3 * K <= N)
  x <- rnorm(N) + 5
  
  group <- sample(1:K, size = N, replace = TRUE)
  # Ensure all groups are actually present at least twice
  group[1:(3*K)] <- rep(1:K, each = 3)

  b_Intercept <- rnorm(1, 5, 1)   
  b_x <- rnorm(1, 0, 1)
  
  sd_group__Intercept <- abs(rnorm(1, 0, 0.75))
  r_group <- matrix(rnorm(K, 0, sd_group__Intercept), 
                 nrow = K, ncol = 1,
                 dimnames = list(1:K, "Intercept"))
  
  sigma <- abs(rnorm(1, 0, 3))
  
  predictor <- b_Intercept + x * b_x + r_group[group]
  y <- rnorm(N, predictor, sigma)
  
  list(
    parameters = list(
      b_Intercept = b_Intercept,
      b_x = b_x,
      sd_group__Intercept = sd_group__Intercept,
      r_group = r_group,
      sigma = sigma
    ),
    generated = data.frame(y = y, x = x, group = group)
  )
}

n_dataset_generator <- SBC_generator_function(one_dataset_generator, N = 25, K = 3)
```

For increased sensitivity, we also add the log likelihood of the data given parameters
as a generated quantity that we'll also monitor.

```{r}
log_lik_gq_func <- generated_quantities(
  log_lik = sum(dnorm(y, b_Intercept + x * b_x + r_group[group], sigma, log = TRUE)))

```


```{r}
set.seed(12239755)
datasets_func <- generate_datasets(n_dataset_generator, 100)
```

```{r}
priors_func <- prior(normal(0,1), class = "b") +
  prior(normal(5,1), class = "Intercept") +
  prior(normal(0,5), class = "sigma") +
  prior(normal(0,0.75), class = "sd")


backend_func <- SBC_backend_brms(y ~ x + (1 | group),  
                            prior = priors_func, chains = 1,
                            template_dataset = datasets_func$generated[[1]])

```


```{r}
results_func <- compute_results(datasets_func, backend_func, thin_ranks = 10, 
                                gen_quants = log_lik_gq_func, 
                    cache_mode = "results", 
                    cache_location = file.path(cache_dir, "func"))
```



```{r results_func_plots}
plot_rank_hist(results_func)
plot_ecdf_diff(results_func)
```

It looks like there is a problem affecting at least the `b_Intercept` and `sigma` parameters.
We may also notice that the `log_lik` (log likelihood derived from all the parameters) is copying
the behaviour of the worst behaving parameter. This tends to be the case in many models, so in models with lots of parameters, it can be useful to add such a term as they make noticing problems easier. There also other advantages as discussed in the `limits_of_SBC` vignette.

What happened is that `brms` by default centers all the predictors, which changes the
numerical values of the intercept (but not other terms). The interaction with the prior than probably also affects the other parameters.

Maybe we don't want `brms` to do this --- using `0 + Intercept` syntax avoids the centering. 


```{r}
# Using 0 + Intercept also changes how we need to specify priors
priors_func2 <- prior(normal(0,1), class = "b") +
  prior(normal(5,1), class = "b", coef = "Intercept") +
  prior(normal(0,5), class = "sigma") +
  prior(normal(0,0.75), class = "sd")


backend_func2 <- SBC_backend_brms(y ~ 0 + Intercept + x + (1 | group),  
                            prior = priors_func2, warmup = 1000, iter = 2000, chains = 1,
                            template_dataset = datasets_func$generated[[1]])


```

Let's fit the same datasets with the new backend.

```{r}
results_func2 <- compute_results(datasets_func, backend_func2, thin_ranks = 10, 
                                 gen_quants = log_lik_gq_func, 
                    cache_mode = "results", 
                    cache_location = file.path(cache_dir, "func2"))
```

We see that this results in mostly non-problematic univariate checks.

TODO: Figure out the source of divergences and check if they create the mismatch for some of the parameters.

```{r results_func2_plots}
plot_rank_hist(results_func2)
plot_ecdf_diff(results_func2)
```




