---
title: "SBC for brms models"
author: "Martin Modrák"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{SBC for brms models}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

This vignette shows how the SBC package supports `brms` models. 
Let's setup the environment:

```{r setup, message=FALSE,warning=FALSE, results="hide"}
library(SBC)
library(brms)
library(ggplot2)

use_cmdstanr <- getOption("SBC.vignettes_cmdstanr", TRUE) # Set to false to use rstan instead

if(use_cmdstanr) {
  options(brms.backend = "cmdstanr")
} else {
  options(brms.backend = "rstan") 
  rstan::rstan_options(auto_write = TRUE)
}

# Using parallel processing
library(future)
plan(multisession)

# The fits are very fast,
# so we force a minimum chunk size to reduce overhead of
# paralellization and decrease computation time.
options(SBC.min_chunk_size = 5)

# Setup caching of results
if(use_cmdstanr) {
  cache_dir <- "./_brms_SBC_cache"
} else { 
  cache_dir <- "./_brms_rstan_SBC_cache"
}
if(!dir.exists(cache_dir)) {
  dir.create(cache_dir)
}

```

## Generating data using `brms`

The `brms` package has a built-in feature to simulate from 
prior corresponding to the model via the `sample_prior = "only"` option. 
This is a bit less useful in model validation
as bug in `brms` (or any mismatch between what `brms` does and what we _think_ it does)
cannot be found as it will most likely affect the generator and the backend in the same way.
Still this can be useful for validating `brms` itself - we'll get to validation
with custom generators in a while. For now, we'll build a generator using `brms` directly.

Generating simulations with this generator requires us to compile a Stan model 
and may thus take a while. Also the exploration is often problematic, so 
to avoid problems, we take a lot of draws and thin the resulting draws heavily.

```{r}
# We need a "template dataset" to let brms build the model.
# The predictor (x) values will be used for data generation,
# the response (y) values will be ignored, but need to be present and 
# of the correct data type
set.seed(213452)
template_data = data.frame(y = rep(0, 15), x = rnorm(15))
priors <- prior(normal(0,1), class = "b") +
  prior(normal(0,1), class = "Intercept") +
  prior(normal(0,1), class = "sigma")
generator <- SBC_generator_brms(y ~ x, data = template_data, prior = priors, 
                                thin = 50, warmup = 10000, refresh = 2000,
                                # Will generate the log density - this is useful, 
                                #but a bit computationally expensive
                                generate_lp = TRUE 
                                )
```



```{r}
set.seed(22133548)
datasets <- generate_datasets(generator, 100)
```

Now we'll build a backend matching the generator (and reuse the compiled model from the
generator)

```{r}
backend <- SBC_backend_brms_from_generator(generator, chains = 1, thin = 1,
                            warmup = 500, iter = 1500,               
                            inits = 0.1)

# More verbose alternative that results in exactly the same backend:
# backend <- SBC_backend_brms(y ~ x, template_data = template_data, prior = priors, warmup = 500, iter = 1000, chains = 1, thin = 1
#                            init = 0.1)
```


Compute the actual results

```{r}
results <- compute_SBC(datasets, backend,
                    cache_mode = "results", 
                    cache_location = file.path(cache_dir, "first"))
```

There are some problems, that we currently choose to ignore (the highest Rhat is barely above
the 1.01 threshold, so it is probably just noise in Rhat computation).

So we can inspect the rank plots. There are no big problems at this resolution.

```{r results_plots}
plot_rank_hist(results)
plot_ecdf_diff(results)
```

## Using custom generator code


Let's take a bit more complex model - with a single varying intercept.

This time we will not use the `brms` model to also simulate from prior, but 
simulate using an R function. This way, we get to learn if `brms` does what we think it does!

Custom generator code also allows us to have different covariate values for each simulation, potentially improving sensitivity
if we want to check the model for a range of potential covariate values.
If on the other hand we are interested in a specific dataset, it might make
more sense to use the predictors as seen in  the dataset in all simulations to focus
our efforts on the dataset at hand.


Let's take a Gaussian model with a single varying intercept.

The data can be generated using the following code - note that we
need to be careful to match the parameter names as `brms` uses them. You
can call `parnames` on a fit to see them.

```{r}
one_sim_generator <- function(N, K) {
  # N - number of datapoints, K number of groups for the varying intercept
  stopifnot(3 * K <= N)
  x <- rnorm(N) + 5
  
  group <- sample(1:K, size = N, replace = TRUE)
  # Ensure all groups are actually present at least twice
  group[1:(3*K)] <- rep(1:K, each = 3)

  b_Intercept <- rnorm(1, 5, 1)   
  b_x <- rnorm(1, 0, 1)
  
  sd_group__Intercept <- abs(rnorm(1, 0, 0.75))
  r_group <- matrix(rnorm(K, 0, sd_group__Intercept), 
                 nrow = K, ncol = 1,
                 dimnames = list(1:K, "Intercept"))
  
  sigma <- abs(rnorm(1, 0, 3))
  
  predictor <- b_Intercept + x * b_x + r_group[group]
  y <- rnorm(N, predictor, sigma)
  
  list(
    variables = list(
      b_Intercept = b_Intercept,
      b_x = b_x,
      sd_group__Intercept = sd_group__Intercept,
      r_group = r_group,
      sigma = sigma
    ),
    generated = data.frame(y = y, x = x, group = group)
  )
}

n_sims_generator <- SBC_generator_function(one_sim_generator, N = 18, K = 5)
```

For increased sensitivity, we also add the log likelihood of the data given parameters
as a derived quantity that we'll also monitor (see the [`limits_of_SBC`](https://hyunjimoon.github.io/SBC/articles/limits_of_SBC.html) 
vignette for discussion on why this is useful).

```{r}
log_lik_dq_func <- derived_quantities(
  log_lik = sum(dnorm(y, b_Intercept + x * b_x + r_group[group], sigma, log = TRUE))
  # Testing CRPS, probably not worth it
  #, CRPS = mean(scoringRules::crps_norm(y, b_Intercept + x * b_x + r_group[group], sigma))
  )

```


```{r}
set.seed(12239755)
datasets_func <- generate_datasets(n_sims_generator, 100)
```

This is then our `brms` backend - note that `brms` requires us to provide a
dataset that it will use to build the model (e.g. to see how many levels of 
various varying intercepts to include):

```{r}
priors_func <- prior(normal(0,1), class = "b") +
  prior(normal(5,1), class = "Intercept") +
  prior(normal(0,5), class = "sigma") +
  prior(normal(0,0.75), class = "sd")


backend_func <- SBC_backend_brms(y ~ x + (1 | group),  
                            prior = priors_func, chains = 1,
                            template_data = datasets_func$generated[[1]])

```
So we can happily compute:

```{r}
results_func <- compute_SBC(datasets_func, backend_func, 
                                dquants = log_lik_dq_func, 
                    cache_mode = "results", 
                    cache_location = file.path(cache_dir, "func"))
```

So that's not looking good! Divergent transitions, Rhat problems... And the
rank plots also show problems:


```{r results_func_plots}
plot_rank_hist(results_func)
plot_ecdf_diff(results_func)
```

It looks like there is a problem affecting at least the `b_Intercept` and `sigma` variables.
We may also notice that the `log_lik` (log likelihood derived from all the parameters) is copying
the behaviour of the worst behaving variable. This tends to be the case in many models, so in models with lots of variables, it can be useful to add such a term as they make noticing problems easier. 

What happened is that `brms` by default centers all the predictors, which changes the
numerical values of the intercept (but not other terms). The interaction with the prior than probably also affects the other variables.

Maybe we don't want `brms` to do this --- using `0 + Intercept` syntax avoids the centering,
so we build a new backend that should match our simulator better


```{r}
# Using 0 + Intercept also changes how we need to specify priors
priors_func2 <- prior(normal(0,1), class = "b") +
  prior(normal(5,1), class = "b", coef = "Intercept") +
  prior(normal(0,5), class = "sigma") +
  prior(normal(0,0.75), class = "sd")


backend_func2 <- SBC_backend_brms(y ~ 0 + Intercept + x + (1 | group),  
                            prior = priors_func2, warmup = 1000, iter = 2000, chains = 1,
                            template_data = datasets_func$generated[[1]])


```

Let's fit the same simulations with the new backend.

```{r}
results_func2 <- compute_SBC(datasets_func, backend_func2, 
                                 dquants = log_lik_dq_func, 
                    cache_mode = "results", 
                    cache_location = file.path(cache_dir, "func2"))
```

We see that this still results in some problematic fits,
but the proportion got lower. At this point I am honestly unsure
what is the issue, but the rank plots look mostly OK:

```{r results_func2_plots}
plot_rank_hist(results_func2)
plot_ecdf_diff(results_func2)
```


I promise to update this vignette once I figure out the source of the problems.
