---
title: "SBC for ADVI and optimizing in Stan (+HMMs)"
author: "Hyunji Moon, Martin Modrák"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{SBC for ADVI and optimizing in Stan (+HMMs)}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---



# Summary 

Computational algorithms such as variational inference (VI) can fail due to the
inability of the approximation family to capture the true posterior, under/over
penalizing tendencies of convergence metric, and slow convergence of the
optimization process. We'll discuss 3 examples: 

- In Example I a simple Poisson model is shown that is
well handled by default ADVI if the size of the data is small, but becomes miscalibrated with larger datasets. It also turns out that for such a simple model using `optimizing` leads to very good results.

- In Example II we discuss a Hidden Markov Model where the approximation by ADVI is imperfect but not very wrong. We also show how the (mis)calibration responds to changing parameters of the ADVI implementation and that `optimizing` performs worse than ADVI.

- In Example III we show that a small modification to the model from Example II  makes the ADVI approximation perform much worse.

When the fit between posterior and approximation family, convergence metric and
its process are checked so that efficiency is gained without sacrificing
accuracy too much, VI can be applied. On top of its role as "the test"
computational algorithms should pass, SBC provides informative inferential results
which directly affect workflow decisions.


# Introduction

HMC can be slow and depending on the joint posterior (as a combination of data,
prior, and likelihood) and the user's goal, deterministic approximation
algorithms can be an aid. To be specific, if the joint posterior is well-formed
enough for reliable approximation (symmetric for ADVI which has normal
approximation family) or the user only needs point estimate (i.e. specification
up to distribution-level is not needed) users can consider the deterministic
alternatives for their inference tool such as ADVI supported by Stan. Note that
Pathfinder (Zhang, 2021) which blends deterministic algorithm's efficiency and
stochastic algorithm's accuracy in a timely manner is under development. SBC
provides one standard to test whether ADVI works well for your model without
ever needing to run full HMC for your model. 

Let's start by setting up our environment. 

```{r setup, message=FALSE,warning=FALSE, results="hide"}
library(SBC)
library(ggplot2)
library(cmdstanr)

library(rstan)
rstan_options(auto_write = TRUE)


options(mc.cores = parallel::detectCores())

# Parallel processing

library(future)
plan(multisession)

# The fits are very fast,
# so we force a minimum chunk size to reduce the overhead of
# paralellization and decrease computation time.
options(SBC.min_chunk_size = 5)


# Setup caching of results
cache_dir <- "./approximate_computation_SBC_cache"
if(!dir.exists(cache_dir)) {
  dir.create(cache_dir)
}

```


# Example I - Poisson 

We'll start by the extremely simple Poisson model already introduced in the 
basic usage vignette:

```{r, comment = ""}
cat(readLines("stan/poisson.stan"), sep = "\n")
```

And here's R code that generates data matching that model:

```{r}
poisson_generator_single <- function(N){  
  # N is the number of data points we are generating
  lambda <- rgamma(n = 1, shape = 15, rate = 5)
  y <- rpois(n = N, lambda = lambda)
  list(
    parameters = list(
      lambda = lambda
    ),
    generated = list(
      N = N,
      y = y
    )
  )
}
```

We'll start with Stan's ADVI with all default parameters, i.e. a mean-field
variational approximation. 
We compile the model and create a variational SBC backend.

```{r}
model_poisson <- cmdstan_model("stan/poisson.stan")
backend_poisson <- SBC_backend_cmdstan_variational(model_poisson, n_retries_init = 3)
```

Note that we allow the backend to retry initialization several times (`n_retries_init`),
as the ADVI implementation in Stan can sometimes fail to start properly on the first try even for very simple models. This ability to retry is an extension in the SBC package and not implemented in Stan.


Throughout the vignette, we'll also use caching for the results.

Since the model runs quickly and is simple, we start with 1000 simulations.

```{r}
set.seed(46522641)
ds_poisson <- generate_datasets(
  SBC_generator_function(poisson_generator_single, N = 20), 
  n_datasets = 1000)
res_poisson <- 
  compute_results(
    ds_poisson, backend_poisson, keep_fits = FALSE,
    cache_mode = "results", cache_location = file.path(cache_dir, "poisson"))
```

Even with the quite high precision afforded by 1000 simulations, the ECDF diff plot and the ranks show no problems - the model is quite well calibrated, although the wavy shape of the ECDF suggest a minor overconfidence of the approximation:

```{r ecdf_rank_poisson}
plot_ecdf_diff(res_poisson)
plot_rank_hist(res_poisson)
```

To put this in different terms we can look at the observed coverage of central 50%, 80% and 95% intervals. We see that the observed coverage for 50% and 80% intervals is a bit lower than expected.

```{r}
empirical_coverage(res_poisson$stats,width = c(0.95, 0.8, 0.5))
```



## Is more data better?

One would expect that the normal approximation implemented in ADVI becomes better with increased size of the data, this is however not necessarily true - let's run the same model, but increase `N` - the number of observed data points:

```{r}
set.seed(23546224)
ds_poisson_100 <- generate_datasets(
  SBC_generator_function(poisson_generator_single, N = 100), 
  n_datasets = 1000)
res_poisson_100 <- 
  compute_results(ds_poisson_100, backend_poisson, keep_fits = FALSE,
                  cache_mode = "results", cache_location = file.path(cache_dir, "poisson_100"))
```

In this case the model becomes clearly overconfident:

```{r ecdf_rank_poisson_100}
plot_ecdf_diff(res_poisson_100)
plot_rank_hist(res_poisson_100)
```

The empirical coverage of the central intervals confirms this:

```{r}
empirical_coverage(res_poisson_100$stats,width = c(0.95, 0.8, 0.5))
```

## Optimizing

If the model is so simple, maybe a simple Laplace approximation around
the posterior mode would suffice? We can use Stan's `optimizing` mode exactly
for that. Although unfortunately, this is currently implemented only in `rstan`
and not for `cmdstanr` (because the underlying CmdStan does not expose the 
Hessian of the optimizing fit).

So let us build an optimizing backend

```{r}
model_poisson_rstan <- stan_model("stan/poisson.stan")
backend_poisson_optimizing <- SBC_backend_rstan_optimizing(model_poisson_rstan)
```

and use it to fit the same datasets - first to the one with `N = 20`.

```{r}
res_poisson_optimizing <- 
  compute_results(ds_poisson, backend_poisson_optimizing, keep_fits = FALSE,
    cache_mode = "results", cache_location = file.path(cache_dir, "poisson_opt"))
```

The resulting ECDF and rank plots are very good.

```{r ecdf_rank_poisson_optimizing}
plot_ecdf_diff(res_poisson_optimizing)
plot_rank_hist(res_poisson_optimizing)
```

Similarly, we can fit the `N = 100` datasets.

```{r}
res_poisson_optimizing_100 <- 
  compute_results(ds_poisson_100, backend_poisson_optimizing, keep_fits = FALSE,
    cache_mode = "results", cache_location = file.path(cache_dir, "poisson_opt_100"))
```

The resulting rank plot once again indicates no serious issues and we thus get
better results here than with ADVI.

```{r ecdf_rank_poisson_optimizing_100}
plot_ecdf_diff(res_poisson_optimizing_100)
plot_rank_hist(res_poisson_optimizing_100)
```

## Summary

We see that for simple models ADVI can provide very tight approximation to exact inference, but this cannot be taken for granted. Surprisingly, having more data does not make the ADVI approximation necessarily better. Additionally, for such simple models, a simple Laplace approximation around the posterior mode works better (and likely faster) than ADVI.

# Example II - Hidden Markov Model

We'll jump to a quite more complex model (partially because we wanted to have a HMM example).

In this example, we have collected a set of counts of particles emitted by 
a specimen in a relatively large number of experimental runs. We however noticed 
that there is a suspiciously large number of low counts. Inspecting the 
equipment, it turns out that the experiment was not set up properly and
in some of the runs, our detector could only register background noise.
We however don't know which runs were erroneous. 

So we assume that some
 experiments
contain both background noise and the signal of interest and the rest
contain just the background. For simplicity, we assume a Poisson distribution
for the counts. 


Additionally, observing background only vs. signal in individual data points is not independent
and we want to model how the experimental setup switches between these two states over
time. We add additional structure to the model to account for this autocorrelation.

One possible choice for such structure is hidden Markov models (HMMs) where
we assume the probability of transitioning from one state to another is identical
across all time points. The [case study for HMMs](https://mc-stan.org/users/documentation/case-studies/hmm-example.html)
has a more thorough discussion and also shows how to code those in Stan.

Maybe the simplest way to describe the model is to show how we simulate the data:

```{r}
generator_HMM <- function(N) {
  
  mu_background <- rlnorm(1, -2, 1)
  mu_signal <- rlnorm(1, 2, 1)

  # Draw the transition probabilities
  t1 <- MCMCpack::rdirichlet(1, c(3, 3))
  t2 <- MCMCpack::rdirichlet(1, c(3, 3))

  states = rep(NA_integer_, N)
  # Draw from initial state distribution
  rho <- MCMCpack::rdirichlet(1, c(1, 10))

  # Simulate the hidden states
  states[1] = sample(1:2, size = 1, prob = rho)
  for(n in 2:length(states)) {
    if(states[n - 1] == 1)
      states[n] = sample(c(1, 2), size = 1, prob = t1)
    else if(states[n - 1] == 2)
      states[n] = sample(c(1, 2), size = 1, prob = t2)
  }  

  # Simulate observations given the state
  mu <- c(mu_background, mu_background + mu_signal)
  y <- rpois(N, mu[states])
  
  list(
    parameters = list(
      mu_background = mu_background,
      mu_signal = mu_signal,
      # rdirichlet returns matrices, convert to 1D vectors
      t1 = as.numeric(t1),
      t2 = as.numeric(t2),
      rho = as.numeric(rho)
    ),
    generated = list(
      N = N,
      y = y
    )
  )
}
```


And here is the Stan code that models this process (it is based on the example
from the HMM case study but simplified and modified).

```{r, comment = ""}
cat(readLines("stan/hmm_poisson.stan"), sep = "\n")
```

## Default ADVI

We start with the default (meanfield) variational backend via Stan:

```{r}
model_HMM <- cmdstan_model("stan/hmm_poisson.stan")
backend_HMM <- SBC_backend_cmdstan_variational(model_HMM, n_retries_init = 3)
```
Since we are feeling confident that our model is implemented correctly (and the model runs quickly), we start with 100 simulations and assume 100 observations for each.
If you are developing a new model, it might be useful to start with fewer simulations, as discussed in the 
[small model workflow vignette](https://hyunjimoon.github.io/SBC/articles/small_model_workflow.html).

And we compute results 

```{r}
set.seed(642354822)
ds_hmm <- generate_datasets(SBC_generator_function(generator_HMM, N = 100), n_datasets = 100)
res_hmm <- compute_results(ds_hmm, backend_HMM,
                           cache_mode = "results", cache_location = file.path(cache_dir, "hmm"))
```

There are not huge problems, but the `mu_signal` variable seems to not be well calibrated:

```{r hmm_ecdf_ranks}
plot_ecdf_diff(res_hmm)
plot_rank_hist(res_hmm)
```

We may also look at the observed coverage of central intervals - we see that for `mu_signal` the approximation tends to be overconfident for the wider intervals.

```{r hmm_coverage}
plot_coverage(res_hmm)
```

To make sure this is not a fluke we add 400 more simulations.

```{r}
set.seed(2254355)
ds_hmm_2 <- generate_datasets(SBC_generator_function(generator_HMM, N = 100), n_datasets = 400)
```

```{r}
res_hmm_2 <- bind_results(
  res_hmm,
  compute_results(ds_hmm_2,backend_HMM,
                  cache_mode = "results",
                  cache_location = file.path(cache_dir, "hmm2"))
)
```

This confirms the problems with `mu_signal`. additionally, we see that `mu_background` and the `rho` parameters also show some irregularities.

```{r hmm_2_ecdf_ranks}
plot_ecdf_diff(res_hmm_2)
plot_rank_hist(res_hmm_2)
```

Looking at the observed coverage, both `mu_background` and `mu_signal` are now clearly somewhat overconfident for the wider intervals.

```{r hmm_2_coverage}
plot_coverage(res_hmm_2)
```




This is what we get when we focus on the 90% posterior credible interval:

```{r}
coverage_hmm <- empirical_coverage(res_hmm_2$stats, width = 0.9)[, c("variable", "ci_low", "ci_high")]
coverage_hmm
```

So the 90% central credible interval for `mu_signal` likely
contains less than `r scales::percent(coverage_hmm$ci_high[2])` of true values.

For a crude result, the default ADVI setup we just tested is not terrible: we don't
expect to see a strong bias and the model will be somewhat overconfident, but
not catastrophically so.

Note that when the user is aiming for a point estimate of mean or other central tendency, a summary of VI posterior may provide a good point estimate even when the uncertainty is miscalibrated. VSBC, a diagnostic that concentrates
on bias in marginal quantity, was developed to test this (Yao et. al., 2018), but is currently not implemented in our package (see https://github.com/hyunjimoon/SBC/issues/60 for progress). Other
diagnostic such as PSIS-based which is associated with specific data and test
quantity, is less flexible for target-testing.

## Full-rank

We may try if the situation improves with full-rank ADVI - let's run it for the same datasets.

```{r}
ds_hmm_all <- bind_datasets(ds_hmm, ds_hmm_2)
res_hmm_fullrank <- compute_results(
  ds_hmm_all, 
  SBC_backend_cmdstan_variational(model_HMM, algorithm = "fullrank", n_retries_init = 3),
  cache_mode = "results", cache_location = file.path(cache_dir, "hmm_fullrank"))
```

We still have problems, but different ones (and arguably somewhat less severe):

```{r hmm_fullrank_ecdf_ranks}
plot_ecdf_diff(res_hmm_fullrank)
plot_rank_hist(res_hmm_fullrank)
```

Interestingly, the rank plot for `mu_signal` shows a "frowning" shape, meaning the
mean-field approximation is slightly underconfident here.

This is nicely demonstrated by looking at the central interval coverage - now the coverage of `mu_signal` is _larger_ than it should be, so the model is underconfident (i.e. more conservative), while the coverages for other parameters track the nominal values quite closely.

```{r hmm_fullrank_coverage}
plot_coverage(res_hmm_fullrank)

```

Or alternatively looking at the numerical values for coverage of the central 90% interval

```{r}
coverage_hmm_fullrank <- 
  empirical_coverage(res_hmm_fullrank$stats, width = 0.9)[, c("variable", "ci_low", "ci_high")]
coverage_hmm_fullrank
```

This pattern where the default meanfield approximation is overconfident and the fullrank approximation is underconfident is in fact quite frequently seen, which
motivated some experiments with a low rank approximation that would fall in between those, but as of early 2022 this is not ready for use in Stan.


## Meanfield + lower tolerance

In some cases, it might also help to reduce the tolerance (`tol_rel_obj`) of the
algorithm. This is a restriction on evidence lower bound (ELBO) for tighter
optimization convergence. Here we'll use the default mean-field
algorithm, but decrease the `tol_rel_obj` (the default value is 0.01). So let's try that.

```{r}
res_hmm_lowtol <- compute_results(
  ds_hmm_all, 
  SBC_backend_cmdstan_variational(model_HMM, tol_rel_obj = 0.001, n_retries_init = 3),
  cache_mode = "results", cache_location = file.path(cache_dir, "hmm_lowtol"))
```

Reducing tolerance leads to a small proportion of non-converging fits. In
theory, increasing `grad_samples` improve non-convergence but in our experience,
current ADVI (2021) convergence does not easily change with this adjustment.
Also, since the non-converged cases are relatively rare, we'll just remove the
non-converging fits from the SBC results (this is OK as long as we would discard
non-converging fits for real data, see the [rejection sampling
vignette](https://hyunjimoon.github.io/SBC/articles/rejection_sampling.html)).

```{r}
res_hmm_lowtol_conv <-
    res_hmm_lowtol[res_hmm_lowtol$backend_diagnostics$elbo_converged] 
```


```{r hmm_lowtol_ecdf_ranks}
plot_ecdf_diff(res_hmm_lowtol_conv)
plot_rank_hist(res_hmm_lowtol_conv)
```

The problems seem to have become even less pronounced. We may once again inspect the
observed coverage of central intervals

```{r hmm_lowtol_coverage}
plot_coverage(res_hmm_lowtol_conv)
```

and the numerical values for the coverage of the central 90% interval.

```{r}
empirical_coverage(res_hmm_lowtol$stats, width = 0.9)[, c("variable", "ci_low", "ci_high")]
```

This variant has somewhat lower overall mismatch, but tends to be overconfident, which might in some cases be less desirable than the more conservative fullrank.

## Optimizing

Would optimizing provide sensible results in this case? We build an optimizng backend and run it.

```{r}
SBC:::require_package_version("rstan", "2.26", "The models in the following sections need more recent rstan than what is available on CRAN - use https://mc-stan.org/r-packages/ to get it")

model_HMM_rstan <- stan_model("stan/hmm_poisson.stan")

res_hmm_optimizing <- compute_results(
  ds_hmm_all, 
  SBC_backend_rstan_optimizing(model_HMM_rstan),
  cache_mode = "results", cache_location = file.path(cache_dir, "hmm_optimizing"))
```





We see that while for some parameters (`mu_signal`, the transition probabilities `t[]`), the Laplace approximation is reasonably well calibrated, it is very
badly calibrated with respect to the initial states `rho` and also for `mu_background`, where there is substantial bias. So if we were only interested in
a subset of the parameters, the optimizing fit could still be on OK choice.

```{r hmm_optimizing_ecdf_ranks}
plot_ecdf_diff(res_hmm_optimizing)
plot_rank_hist(res_hmm_optimizing)
```

## Summary

To summarise, the HMM model turns out to pose minor problems for ADVI that can be partially resolved by tweaking the parameters of the ADVI algorithm. Just using optimizing results in much worse calibration than ADVI.

Another relevant question is how much speed we gained. To have a comparison, we run full MCMC with Stan for the same datasets.

```{r}
res_hmm_sample <- compute_results(
  ds_hmm[1:50], 
  SBC_backend_cmdstan_sample(model_HMM),
  keep_fits = FALSE,
  cache_mode = "results", cache_location = file.path(cache_dir, "hmm_sample"))
```

We get a small number of problematic fits, which we will ignore for now. We check that there are no obvious calibration problems:

```{r hmm_sample_ecdf_ranks}
plot_ecdf_diff(res_hmm_sample)
plot_rank_hist(res_hmm_sample)
```

For the machine we built the vignette on, here are the distributions
of times (for ADVI and optimizing) and time of longest chain (for HMC):

```{r hmm_time}
hmm_time <- 
  rbind(
      data.frame(alg = "Optimizing", 
                     time = res_hmm_optimizing$backend_diagnostics$time),
      data.frame(alg = "Meanfield", 
                     time = res_hmm$backend_diagnostics$time),
        data.frame(alg = "Fullrank", 
                   time = res_hmm_fullrank$backend_diagnostics$time),
        data.frame(alg = "Meanfield + low tol.", 
                   time = res_hmm_lowtol$backend_diagnostics$time),
        data.frame(alg = "Sampling (longest chain)", 
                   time = res_hmm_sample$backend_diagnostics$max_chain_time))

max_time_optimizing <- round(max(res_hmm_optimizing$backend_diagnostics$time), 2)


hmm_time$alg <- factor(hmm_time$alg, 
                       levels = c("Optimizing",
                                  "Meanfield",
                                  "Fullrank",
                                  "Meanfield + low tol.",
                                  "Sampling (longest chain)"))
 
ggplot(hmm_time, aes(x = time)) + 
  geom_histogram(aes(y = ..density..), bins = 20) + 
  facet_wrap(~alg, ncol = 1) +
  scale_x_continuous("Time [seconds]")
```

Depressingly, while using lower tolerance let us get almost as good uncertainty
quantification as sampling, it also erased a big part of the performance advantage variational
inference had over sampling for this model. However, both the fullrank and meanfield approximations provide not-terrible estimates and are noticeably faster than sampling. Optimizing is by far the fastest as the longest time observed is just `r max_time_optimizing` seconds.



# Example III - Hidden Markov Model, ordered variant

Unforutnately, ADVI as implemented in Stan can be quite fragile. Let us consider
a very small change to the HMM model from the previous section: let us model the means of the counts for the two states directly (the previous version modelled the background state and the difference between the two states) and move to the log scale.
So instead of `mu_background` and `mu_signal` we have an `ordered` vector `log_mu`:


```{r, comment = ""}
cat(readLines("stan/hmm_poisson_ordered.stan"), sep = "\n")
```

This model is almost identical - in theory the only difference is that it implies a slightly different prior on the active (higher mean) state. Here is how we can
generate data with this mildly different prior (we need rejection sampling to fulfill the ordering constraint):

```{r}
generator_HMM_ordered <- function(N) {
  
  # Rejection sampling for ordered mu with the correct priors
  repeat {
    log_mu <- c(rnorm(1, -2, 1), rnorm(1, 2, 1))
    if(log_mu[1] < log_mu[2]) {
      break;
    }
  }
  
  mu <- exp(log_mu)

  # Draw the transition probabilities
  t1 <- MCMCpack::rdirichlet(1, c(3, 3))
  t2 <- MCMCpack::rdirichlet(1, c(3, 3))

  states = rep(NA_integer_, N)
  # Draw from initial state distribution
  rho <- MCMCpack::rdirichlet(1, c(1, 10))

  states[1] = sample(1:2, size = 1, prob = rho)
  for(n in 2:length(states)) {
    if(states[n - 1] == 1)
      states[n] = sample(c(1, 2), size = 1, prob = t1)
    else if(states[n - 1] == 2)
      states[n] = sample(c(1, 2), size = 1, prob = t2)
  }  
  
  y <- rpois(N, mu[states])
  
  list(
    parameters = list(
      log_mu = log_mu,
      # rdirichlet returns matrices, convert to 1D vectors
      t1 = as.numeric(t1),
      t2 = as.numeric(t2),
      rho = as.numeric(rho)
    ),
    generated = list(
      N = N,
      y = y
    )
  )
}
```


So let us build a default variational backend and fit it to just 20 datasets.

```{r}
model_HMM_ordered <- cmdstan_model("stan/hmm_poisson_ordered.stan")
backend_HMM_ordered <- SBC_backend_cmdstan_variational(model_HMM_ordered, n_retries_init = 3)
```


```{r}
set.seed(12333654)
ds_hmm_ordered <- generate_datasets(
  SBC_generator_function(generator_HMM_ordered, N = 100), 
  n_datasets = 20)

res_hmm_ordered <- 
  compute_results(ds_hmm_ordered, backend_HMM_ordered,
                  cache_mode = "results", cache_location = file.path(cache_dir, "hmm_ordered"))
```


Immediately we see that the `log_mu[1]` parameter is heavily miscalibrated.

```{r hmm_ordered_ecdf_ranks}
plot_ecdf_diff(res_hmm_ordered)
plot_rank_hist(res_hmm_ordered)
```

What changed? To understand that we need to remember how
Stan [represents constrained data types]( https://mc-stan.org/docs/2_28/reference-manual/variable-transforms.html). In short,
in the model in Example II, Stan will internally work with the so called _unconstrained_ parameters `mu_background__ = log(mu_background)` and 
`mu_signal__ = log(mu_signal)`. In this modified model, the internal representation will be: `log_mu_1__ = log_mu[1]` (without any change) and `log_mu_2__ = log(log_mu[2] - log_mu[1])`. So the mean for the active component is actually `exp(log_mu_1__ + exp(log_mu_2__))`. This then introduces a complex correlation structure between the unconstrained parameters that the ADVI algorithm is unable to handle well.


Even trying the fullrank variant does not help:

```{r}
backend_HMM_ordered_fullrank <- 
  SBC_backend_cmdstan_variational(model_HMM_ordered,
                                  algorithm = "fullrank", n_retries_init = 3)

res_hmm_ordered_fullrank <- 
  compute_results(ds_hmm_ordered, backend_HMM_ordered,
                  cache_mode = "results", cache_location = file.path(cache_dir, "hmm_ordered_fullrank"))
```


The results are still strongly miscalibrated.

```{r hmm_ordered_fullrank_ecdf_ranks}
plot_ecdf_diff(res_hmm_ordered_fullrank)
plot_rank_hist(res_hmm_ordered_fullrank)
```

To have a complete overview we may also try the optimizing fit:

```{r}
model_HMM_ordered_rstan <- stan_model("stan/hmm_poisson_ordered.stan")

res_hmm_ordered_optimizing <- compute_results(
  ds_hmm_ordered, 
  SBC_backend_rstan_optimizing(model_HMM_ordered_rstan),
  cache_mode = "results", cache_location = file.path(cache_dir, "hmm_ordered_optimizing"))
```

in this case, optimizing has better calibration for `log_mu`, but worse calibration for `rho` than ADVI.

```{r hmm_ordered_optimizing_ecdf_ranks}
plot_ecdf_diff(res_hmm_ordered_optimizing)
plot_rank_hist(res_hmm_ordered_optimizing)
```

# Conclusion

As this vignette has shown, for some models, ADVI will provide results that are 
close to what we get with sampling, but it may also fail catastrophically on models that are just slightly different. Tweaking the algorithm parameters might also be necessary. For some cases where ADVI works, the Laplace approximation with optimizing will also work well. ADVI (and optimizng) cannot thus be used blindly. Fortunately SBC can be used to check against this type of problem without ever needing to run the full sampling.

# Next step: Evolving computation and diagnostic.

In computational_algorithm2, we will focus on hopeful aspects of approximate
computation. The adversarial relation between computation and diagnostic is
introduced based on which mutual evolvement happens. This can give insight to
computational algorithm designers aiming to pass SBC. For illustration, when and
how VI can be used is discussed which include customized SBC (e.g. VSBC) and first or
second-order correction.

# References
- Zhang et. al. (2021) Pathfinder: Parallel quasi-Newton variational inference https://arxiv.org/abs/2108.03782
- Turner and Sahani (2010) Two problems with variational expectation maximisation for
time-series models http://www.gatsby.ucl.ac.uk/~maneesh/papers/turner-sahani-2010-ildn.pdf
- Yao et. al. (2018) Yes, but Did It Work?: Evaluating Variational Inference http://proceedings.mlr.press/v80/yao18a/yao18a.pdf
