---
title: "SBC with discrete parameters in Stan and JAGS"
author: "Martin Modrák"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{SBC with discrete parameters in Stan and JAGS}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

SBC was primarily designed for continuous parameters, but can be used
with models that have discrete parameters - whether the parameters
are directly represented (e.g. in BUGS/JAGS) or marginalized out (as is usual in Stan).

## Stan version and debugging

```{r setup, message=FALSE,warning=FALSE, results="hide"}
library(SBC); 
library(ggplot2)

use_cmdstanr <- getOption("SBC.vignettes_cmdstanr", TRUE) # Set to false to use rstan instead

if(use_cmdstanr) {
  library(cmdstanr)
} else {
  library(rstan)
  rstan_options(auto_write = TRUE)
}

# Multiprocessing support
library(future)
plan(multisession)

# The fits are very fast and we fit just a few, 
# so we force a minimum chunk size to reduce overhead of
# paralellization and decrease computation time.
options(SBC.min_chunk_size = 5)

# Setup caching of results
if(use_cmdstanr) {
  cache_dir <- "./_discrete_vars_SBC_cache"
} else {
  cache_dir <- "./_discrete_vars_rstan_SBC_cache"
}
cache_dir_jags <- "./_discrete_vars_SBC_cache"

if(!dir.exists(cache_dir)) {
  dir.create(cache_dir)
}
if(!dir.exists(cache_dir_jags)) {
  dir.create(cache_dir_jags)
}

```

We take the changepoint model from:
https://mc-stan.org/docs/2_26/stan-users-guide/change-point-section.html

```{r, comment = ""}
cat(readLines("stan/discrete_vars1.stan"), sep = "\n")
```

```{r}
if(use_cmdstanr) {
  model_1 <- cmdstan_model("stan/discrete_vars1.stan")
  backend_1 <- SBC_backend_cmdstan_sample(model_1)
} else {
  model_1 <- stan_model("stan/discrete_vars1.stan")
  backend_1 <- SBC_backend_rstan_sample(model_1)
}
```

Now, let's generate data from the model.

```{r}
generate_single_sim_1 <- function(T, r_e, r_l) {
  e <- rexp(1, r_e)
  l <- rexp(1, r_l)
  s <- sample.int(T, size = 1)
  
  y <- array(NA_real_, T)
  for(t in 1:T) {
    if(t <= s) {
      rate <- e
    } else {
      rate <- l
    }
    y[t] <- rpois(1, rate) 
  }
  
  list(
    variables = list(
      e = e, l = l, s = s
    ), generated = list(
      T = T,
      r_e = r_e,
      r_l = r_l,
      y = y
    )
  )
}

generator_1 <- SBC_generator_function(generate_single_sim_1, T = 5, r_e = 0.5, r_l = 0.1)
```


```{r}
set.seed(85394672)
datasets_1 <- generate_datasets(generator_1, 30)

```

Additionally,
we'll add a generated quantity expressing the total log-likelihood of data given the
fitted parameters. The expression within the `generated_quantities()` call is evaluated
for both prior and posterior draws and
included as another variable in SBC checks. It turns out this type of generated quantities
can increase the sensitivity of the SBC against some issues in the model. 
See `vignette("limits_of_SBC")` for a more detailed discussion of this.

```{r}
log_lik_gq <- generated_quantities(log_lik = sum(dpois(y, ifelse(1:T < s, e, l), log = TRUE)))
```

So finally, lets actually compute SBC:

```{r}
results_1 <- compute_SBC(datasets_1, backend_1, 
                    cache_mode = "results", 
                    cache_location = file.path(cache_dir, "model1"),
                    gen_quants = log_lik_gq)
```

Here we also use the caching feature to avoid recomputing the fits when recompiling this vignette. 
In practice, caching is not necessary but is often useful.

TODO the diagnostic failures are false positives, because Rhat and ESS don't work very well for discrete parameters.
We need to figure out how to handle this better.

We can quickly note that the statistics for the `s` parameter are extreme - many ranks of 0 and _extreme_ z-scores, including -Infinity. Seing just one or two such fits should be enough to convince us that there is something fundamentally wrong.

```{r}
dplyr::filter(results_1$stats, variable == "s") 
```


Inspecting the statistics shows that quite often, the model is quite sure of the value of `s` while the simulated value is just one less. 

Looking at the `ecdf_diff` plot we see that this seems to compromise heavily the inference for `s`, but the other parameters do not show such bad behaviour. Note that the `log_lik` generated quantity shows even starker
failure than `s`, so it indeed poses a stricter check in this scenario. 

```{r results1_plots}
plot_ecdf_diff(results_1)
plot_rank_hist(results_1)
```

An important note: you may wonder, how we got such a wiggly line for the `s` parameter - doesn't it
have just 5 possible values? Shouldn't therefore the ECDF be one big staircase? 
In fact the package does a little trick to make discrete parameters comparable to
continuous - the rank of a discrete parameter is chosen uniformly randomly across
all possible ranks (i.e. posterior draws that have exactly equal value). This 
means that if the model is well behaved, ranks for the discrete parameter will
be uniformly distributed across the whole range of possible ranks and we can
use exactly the same diagnostics for a discrete parameter as we do for the 
continuous ones.

But back to the model - what happened? What is wrong with it? After some inspection, you may notice that the simulator does not match the model - the model takes the early rate (`e`) for points `t < s` while the simulator takes `e` for points `t <=  s`, so there is effectively a shift by one time point between the simulator and the model. So let's assume that we beleive that the Stan model is in fact right. We therefore updated the simulator to match the model:


```{r}
generate_single_sim_2 <- function(T, r_e, r_l) {
  e <- rexp(1, r_e)
  l <- rexp(1, r_l)
  s <- sample.int(T, size = 1)
  
  y <- array(NA_real_, T)
  for(t in 1:T) {
    if(t < s) { ### <--- Only change here
      rate <- e
    } else {
      rate <- l
    }
    y[t] <- rpois(1, rate) 
  }
  
  list(
    variables = list(
      e = e, l = l, s = s
    ), generated = list(
      T = T,
      r_e = r_e,
      r_l = r_l,
      y = y
    )
  )
}

generator_2 <- SBC_generator_function(generate_single_sim_2, T = 5, r_e = 0.5, r_l = 0.1)
```

And we can recompute:


```{r}
set.seed(5846502)
datasets_2 <- generate_datasets(generator_2, 30)
results_2 <- compute_SBC(datasets_2, backend_1,
                    gen_quants = log_lik_gq, 
                    cache_mode = "results", 
                    cache_location = file.path(cache_dir, "model2"))
```

```{r results_2_plots}
plot_rank_hist(results_2)
plot_ecdf_diff(results_2)
```


Looks good, so let us add some more simulations to make sure the model behaves well.

```{r}
set.seed(54321488)
datasets_2_more <- generate_datasets(generator_2, 100)
results_2_more <- compute_SBC(datasets_2_more, backend_1,
                    gen_quants = log_lik_gq, 
                    cache_mode = "results", 
                    cache_location = file.path(cache_dir, "model3"))

results_2_all <- bind_results(results_2, results_2_more)
```

```{r results_2_all_plots}
plot_rank_hist(results_2_all)
plot_ecdf_diff(results_2_all)
```


Now - as far as this amount of SBC steps can see, the model is good and we get good behaviour for both the continuous and the discrete parameters and the `log_lik` generated quantity. Hooray!

## JAGS version

We can now write the same model in JAGS. This becomes a bit easier as JAGS lets
us represent discrete parameters directly:

```{r, comment = ""}
cat(readLines("other_models/changepoint.jags"), sep = "\n")
```

We will use the `rjags` package, let us verify it is installed correctly.


```{r}
library(rjags)
```


We will also default to relatively large number of samples as 
we can expect some autocorrelation in the Gibbs sampler.

```{r}
backend_jags <- SBC_backend_rjags("other_models/changepoint.jags",
                                  variable.names = c("e","l","s"),
                                  n.iter = 10000,
                                  n.burnin = 1000,
                                  n.chains = 4,
                                  thin = 10)
```


Running SBC with all the corrected datasets from before (rJAGS accepts input
data in exactly the same format as Stan, so we can reuse the datasets without any change):

```{r}
datasets_2_all <- bind_datasets(datasets_2, datasets_2_more)
results_jags <- compute_SBC(datasets_2_all, backend_jags,
                            gen_quants = log_lik_gq,
                        cache_mode = "results",
                        cache_location = file.path(cache_dir_jags, "rjags"))
```

Similarly to the case above, the Rhat and ESS warnings are false positives due to the `s` parameter, which
we need to handle better.

The rank plots show no problems.

```{r ranks_jags}
plot_rank_hist(results_jags)
plot_ecdf_diff(results_jags)
```

As an exercise, we can also write the marginalized version of the model in JAGS.
In some cases, marginalization improves performance even for JAGS models, however,
for this model it is actually not an improvement, presumably because the model
is very simple.


```{r, comment = ""}
cat(readLines("other_models/changepoint_marginalized.jags"), sep = "\n")
```

The code got quite a bit more complex, se let's check if we didn't mess up
the rewrite - first we 
build a backend with this new representation:

```{r}
backend_jags_marginalized <- SBC_backend_rjags("other_models/changepoint_marginalized.jags",
                                  variable.names = c("e","l","s"),
                                  n.iter = 10000,
                                  n.burnin = 1000,
                                  n.chains = 4,
                                  thin = 10)
```

Then we run the actual SBC:

```{r}
results_jags_marginalized <- compute_SBC(datasets_2_all, backend_jags_marginalized,
                                         gen_quants = log_lik_gq,
                        cache_mode = "results",
                        cache_location = file.path(cache_dir_jags, "rjags_marginalized"))
```

And the ranks plots look good, so we indeed probably did succeed in correctly marginalizing
the `s` parameter!

```{r  ranks_jags_marginalized}
plot_rank_hist(results_jags_marginalized)
plot_ecdf_diff(results_jags_marginalized)
```



