---
title: "Implementing a new backend (algorithm) + frequentist approximations"
author: "Martin Modrák"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Implementing a new backend (algorithm) + frequentist approximations}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

This vignette will discuss how to implement a new type of backend for the SBC package
and thus allow you to integrate the SBC package with any method/algorithm that
can produce draws from a posterior distribution or its approximation.
As an example, we'll wrap the base R `glm` function as a backend. This will
also let us discuss how we can treat frequentist models as approximations to
Bayesian models and that SBC can tell us how faithful such an approximation is.

We assume familiarity with the SBC package architecture as discussed in the  [basic usage vignette](https://hyunjimoon.github.io/SBC/articles/basic_usage.html) and
[S3 classes](https://adv-r.hadley.nz/s3.html) in R.

Let's setup the environment.

```{r setup, message=FALSE,warning=FALSE, results="hide"}
library(SBC)
library(ggplot2)
library(medicaldata)
library(formula.tools)
library(MASS)

# Setup caching of results
cache_dir <- "./_implementing_backends_SBC_cache"
if(!dir.exists(cache_dir)) {
  dir.create(cache_dir)
}

```

## Minimal backend support

If you remember from the, [interface introduction vignette](https://hyunjimoon.github.io/SBC/articles/basic_usage.html)
a backend for the SBC package describes a function that takes in data and produces posterior draws,
i.e. the backend holds all the information other than data that are needed to 
run the given statistical method. 

For practical reasons, the SBC package actually splits that function into two steps: first,
there is an S3 generic [`SBC_fit()`](https://hyunjimoon.github.io/SBC/reference/SBC_fit.html),
that takes a backend object, observed data and the number of cores it is allowed to use
and produces an arbitrary object representing the fit. Additionally,
there is an `SBC_fit_to_draws_matrix()` S3 generic that takes in the resulting fit
and returns posterior draws in the `posterior::draws_matrix` format. 
The split here is useful because it lets the `SBC_results` object to store the
raw fit objects, which can then be inspected by user for debugging purposes.
The SBC package
makes no assumptions on the fit objects beyond the fact that they define an `SBC_fit_to_draws_matrix` method.
To be precise, even this is not necessary, because if the object implements a method for `as_draws_matrix`
but not `SBC_fit_to_draws_matrix`, the `as_draws_matrix` implementation will be called.

So a simple implementation wrapping the `glm` function will consist of three
short functions. First is a "constructor" function that creates new instance
of the backend. Here, we'll just capture all the arguments (which we will later pass
 to `glm`), the created object will be of a new S3 class `SBC_backend_glm`:

```{r}
SBC_backend_glm <- function(...) {
    args = list(...)
    if(any(names(args) == "data")) {
      stop(paste0("Argument 'data' cannot be provided when defining a backend",
                  " as it needs to be set by the SBC package"))
    }
  
    structure(list(args = args), class = "SBC_backend_glm")
}
```

So e.g. `SBC_backend_glm(y ~ x, family = "poisson")` would create a valid
backend representing a simple Poisson regression.

Now we create an implementation of `SBC_fit` for the newly created class.
We take the generated data (`generated` argument) and pass it - along
with all the arguments we stored in the constructor - to `glm` via `do.call`. 
We ignore the `cores` argument as we don't have multicore support.

```{r}
SBC_fit.SBC_backend_glm <- function(backend, generated, cores) {
  args_with_data <- backend$args
  args_with_data$data <- generated
  
  do.call(glm, args_with_data)
}
```

In some cases, it might be undesirable to work with the raw fit as returned by the underlying
package and wrap the result in some custom class, but here we'll have no trouble 
working with the `glm` S3 class (as returned by the `glm()` function) directly.

The most conceptually difficult part is then the implementation of `SBC_fit_to_draws_matrix` 
for the `glm` S3 class. Here, we'll remember how actually `glm` fits the model:
it finds the maximum likelihood estimate (MLE) and then uses the Hessian
to construct
a multivariate normal approximation to the likelihood around the MLE.
In this way we can see `glm` as an approximate Bayesian method where:

- A flat improper prior is used for all parameters
- The posterior is approximated by a multivariate normal distribution

And that's exactly what we'll do: the `coef` method for `glm` fit returns the
MLE and the `vcov` method returns the variance-covariance matrix implied by the
Hessian, so all we need is to take a bunch of draws (here 1000) from this
multivariate normal. Therefore, the implementation is also very simple:

```{r}
SBC_fit_to_draws_matrix.glm <- function(fit) {
  samp_matrix <- MASS::mvrnorm(n = 1000, mu = coef(fit), Sigma = vcov(fit))
  posterior::as_draws_matrix(samp_matrix)
}
```

Note that for non-base packages, we specify the namespace explicitly and
do not rely on those functions being loaded via `library` - this is good
practice for package development and will make paralellization work (see notes below).

A quick example to show this minimal setup already works. We'll build a simple Poisson
regression simulator:

```{r}
generator_single_poisson <- function(N) {
  log_intercept <- rnorm(1, mean = 4,  sd = 1.5)
  beta <- rnorm(1, mean = 0, sd = 1)
  X <- rnorm(N, mean = 0, sd = 1)
  mus <- log_intercept + beta * X
  y <- rpois(N, exp(mus))
  
  list(
    variables = list(
      # Naming the variables in the same way glm will name coefs
      `(Intercept)` = log_intercept,
      x = beta
    ),
    generated = data.frame(y = y, x = X)
  )
}

set.seed(354662)
datasets_poisson <- generate_datasets(SBC_generator_function(generator_single_poisson, N = 100), 
                              n_sims = 100)
```

Then we'll construct a matching backend and compute the results.

```{r}
backend_poisson <- SBC_backend_glm(y ~ x, family = "poisson")
res_poisson <- compute_SBC(datasets_poisson, 
                               backend_poisson, 
                               thin_ranks = 1,
                               cache_mode = "results", 
                               cache_location = file.path(cache_dir,"poisson"))
```


We have set `thin_ranks = 1` as no thinning is needed (the draws are i.i.d. by
construction).

The rank and ecdf plots show no big problems

```{r poisson_ranks}
plot_rank_hist(res_poisson)
plot_ecdf_diff(res_poisson)
```

This is not unexpected - we've used a large number of observations and a simple model, so
choice of prior should have negligible impact on the posterior and the normal
approximation is very close to the exact Bayesian solution. 

We can see that both variables are recovered almost exactly in almost all 
fits:

```{r poisson_sim_estimated}
plot_sim_estimated(res_poisson)
```


## Additional backend improvements

Beyond the minimal setup, there are additional functions that our backend can implement to make it more
comfortable to use. Let's walk through the options and se and how they can be implemented for the 
`glm` wrapper.

Since (unlike MCMC methods) the `glm` approximation does 
not produce autocorrelated draws, we can implement `SBC_backend_iid_draws`
to return `TRUE`. The SBC package will then by default use `thin_ranks = 1`
argument to `compute_SBC` and will not assess convergence/autocorrelation via the
R-hat and ESS diagnostics.

```{r}
SBC_backend_iid_draws.SBC_backend_glm <- function(backend) {
  TRUE
}
```

Backends based on MCMC may want to implement `SBC_backend_default_thin_ranks()`
which specifies the default thinning needed to remove autocorrelation from the fit.

Most statistical algorithms also give us some diagnostics that let us
understand whether there was some problem in fitting.

To see some of the problems that `glm` can encounter, 
we'll run a quite pathological logistic regression:

```{r}
problematic_data <- data.frame(y = rep(0:1, each = 100), x = 1:200)
glm(y ~ x, data = problematic_data, family = "binomial")
```
So we get two problems - one is that the optimization method did not converge
(this can be checked via the `$converged` member of the resulting fit) and
that extreme probabilities occurred (this is related to the problem of "perfect 
separation"). This problem is however only signalled as a warning and never 
reported in the fit object itself. 

For this reason, SBC package will capture all output, warnings and messages
that the backend produced and our implementation can then inspect and process
all of the output. This can be achieved by implementing the `SBC_fit_to_diagnostics()`
generic for our fit object. This method should return a single row `data.frame`
that contains any diagnostics. Additionally, it is often useful to add
a custom class to the results to allow for automatic summaries (we'll get there).
This is our implementation for the `glm` class:


```{r}
SBC_fit_to_diagnostics.glm <- function(fit, fit_output, fit_messages, fit_warnings) {
  res <- data.frame(
    probs_0_1 = any(grepl("fitted probabilities numerically 0 or 1 occurred", fit_warnings)),
    converged = fit$converged
  )

  class(res) <- c("SBC_glm_diagnostics", class(res))
  res
}
```

Having a custom class let's us implement a `summary` implementation for our
diagnostics:

```{r}
summary.SBC_glm_diagnostics <- function(x) {
  summ <- list(
    n_fits = nrow(x),
    n_probs_0_1 = sum(x$probs_0_1),
    n_not_converged = sum(!x$converged)
  )

  structure(summ, class = "SBC_glm_diagnostics_summary")  
}
```

and we can then implement the `get_diagnostic_messages()` generic, which
takes an object and returns a list of messages potentially signalling problems -
any messages that signal problems are then reported automatically to the user.
The OK messages are also reported when calling `summary` on an `SBC_results` object.

We'll use our `summary` implementation for `SBC_glm_diagnostics` to create the 
messages:

```{r}
get_diagnostic_messages.SBC_glm_diagnostics <- function(x) {
  get_diagnostic_messages(summary(x))
}


get_diagnostic_messages.SBC_glm_diagnostics_summary <- function(x) {
  message_list <- list()
  
  if(x$n_probs_0_1 == 0) {
    message_list[[1]] <- 
      data.frame(ok = TRUE, message = "No fit had 0/1 probabilities.")
  } else {
    message_list[[1]] <- 
      data.frame(ok = FALSE, 
                 message = paste0(
                   x$n_probs_0_1, " (", round(100 * x$n_probs_0_1 / x$n_fits), 
                   "%) of fits had 0/1 probabilities."))    
  }
  
  if(x$n_not_converged == 0) {
    message_list[[2]] <- 
      data.frame(ok = TRUE, message = "All fits converged.")
  } else {
    message_list[[2]] <- 
      data.frame(ok = FALSE, 
                 message = paste0(
                   x$n_not_converged, " (", round(100 * x$n_not_converged / x$n_fits), 
                   "%) of fits did not converge."))    
  }  
  
  SBC_diagnostic_messages(do.call(rbind, message_list))
}
```

Finally, some backend objects are complex and may differ between R sessions,
even if they represent the same backend (Stan backends are a prime example as
the pointer/path to a compiled model can change). If that's the case, it may
break the built-in caching functionality. Such backends may want to implement the
`SBC_backend_hash_for_cache()` generic to provide a hash with better properties.

And that's about all that we can do for our backend. Before we'll use
the extended backend for some investigations, we'll make an aside
about parallel support.

## Considerations for parallelization

SBC uses the `future` package to allow paralellization. This means that
when user sets up a parallel environment (e.g. via `plan(multisession)`),
the `SBC_fit()`, `SBC_fit_to_draws_matrix()`  and `SBC_backend_iid_draws()`
implementations will run in a fresh session. To make this work smoothly,
the functions should call non-base R functions explicitly via namespace declaration 
(e.g. note that we call `MASS::mvrnorm`, not just `mvrnorm`).

If you are implementing the backend to become part of the SBC package, nothing
more is needed for paralellization to work. 
If however you are just building an ad-hoc backend that lives
in your global environment, you will also need to pass the three functions 
to the `globals` argument of `compute_SBC` which will make them available on 
all workers i.e. use:

```
compute_SBC(..., globals = c("SBC_fit.SBC_backend_glm", 
                                 "SBC_fit_to_draws_matrix.glm",
                                 "SBC_backend_iid_draws.SBC_backend_glm"))

```

If those functions in turn depend on other functions not defined in a package,
those functions would need to be added to `globals` as well. In some future
version of the package we hope to be able to autodetect those dependencies.

Also note that if you are OK with single-core processing (which with `glm` is 
very fast), you don't need to care about any of this.

## `glm` as approximate Bayesian method

As we discussed earlier, we can view `glm` as an approximation to a fully Bayesian
method where the posterior is approximate and priors are flat. 
We would therefore expect the approximation to be well-behaved when we
use wide priors for simulation and/or have a lot of data to inform all the 
model coefficients. The obvious way to verify whether the approximation
is good is to run both full Bayesian inference (e.g. via `rstanarm`) and `glm` on the same data. 
The problem is that this reduces the appeal of the approximation, as we need to
wait for the fully Bayesian fit anyway and so we might as well just use the 
Bayesian version.

However, SBC allows another way - we can check that the approximation is good
by running only the approximate method (a lot of times) and look at SBC results.
This may still be faster than running a single fully Bayesian fit. Additionally,
fitting with an approximate algorithm can be useful to run 
approximate power calculations where it lets us cheaply fit a lot of
simulations to e.g. understand how
the width of our posterior intervals changes with sample size
and at the same time we learn, whether the approximation 
is problematic in some way.

For the sake of example, let's assume we've already gathered data that
we want to analyze with Bayesian logistic regression. So our data generating
process will use the observed covariate values but simulate new coefficients
and outcome data. Below is a simple implementation with normal priors
on the intercept and predictors. Note that we do some [rejection
sampling](https://hyunjimoon.github.io/SBC/articles/rejection_sampling.html) 
here to avoid using simulations where the generated response
is the same or almost the same for all rows.

```{r}
generator_single_logistic <- function(formula, 
                                      template_data, 
                                      intercept_prior_loc = 0,
                                      intercept_prior_width = 2,
                                      predictor_prior_loc = 0,
                                      predictor_prior_width = 1) {
  response_name <- formula.tools::lhs.vars(formula)
  if(length(response_name) != 1) {
    stop("The formula has to have just a single response")
  }
  
  X <- model.matrix(formula, template_data)
  
  repeat {
    coefs <- rnorm(ncol(X), predictor_prior_loc, sd = predictor_prior_width)
    names(coefs) <- colnames(X)
    if("(Intercept)" %in% names(coefs)) {
      coefs["(Intercept)"] <- rnorm(1, intercept_prior_loc, sd = intercept_prior_width)
    }
    
    linpred <- X %*% coefs
    probs <- plogis(linpred)
    y <- rbinom(nrow(X), size = 1, prob = probs)
    if(sum(y == 0) >= 5 && sum(y == 1) >= 5) {
      break;
    }
  }

  data_mod <- template_data
  data_mod[[response_name]] <- y

  list(
    variables = as.list(coefs),
    generated = data_mod
  ) 
}
```


We are going to use the `indo_rct` dataset from the `medicaldata` package as a template. The 
dataset contains the results of a randomized, placebo-controlled, 
prospective 2-arm trial of indomethacin 100 mg PR once vs. placebo to prevent post-ERCP Pancreatitis in 602 patients. You can inspect the [codebook](https://htmlpreview.github.io/?https://github.com/higgi13425/medicaldata/blob/master/man/codebooks/indo_rct_codebook.html) as well as the [published paper](https://www.nejm.org/doi/full/10.1056/NEJMoa1111103)
online. The citation for the paper is:

Elmunzer, Higgins, et al., A Randomized Trial of Rectal Indomethacin to Prevent Post-ERCP Pancreatitis, New England Journal of Medicine, 2012, volume 366, pages 1414-1422, as found [here](https://www.nejm.org/doi/full/10.1056/NEJMoa1111103).

### Well-informed model

We'll start by testing our approximate computation with the simplest analysis - the primary binary outcome predicted only
by the treatment (the `rx` column in the data):

```{r}
formula_indo_simple <- outcome ~ rx
set.seed(6524243)
datasets_indo_simple <- generate_datasets(SBC_generator_function(
  generator_single_logistic, 
  formula = formula_indo_simple,
  template_data = medicaldata::indo_rct),
  n_sims = 500) 

backend_indo_simple <- SBC_backend_glm(formula = formula_indo_simple, family = "binomial") 
```


```{r}
res_indo_simple <- compute_SBC(datasets_indo_simple, backend_indo_simple,
                                   cache_mode = "results", 
                                   cache_location = file.path(cache_dir,"indo_simple"))
```

The rank plots look good:

```{r indo_simple_ranks}
plot_rank_hist(res_indo_simple)
plot_ecdf_diff(res_indo_simple)
```

The coverages are very tight

```{r indo_sim_coverage}
plot_coverage(res_indo_simple)
```

we can make this precise by inspecting the same results numerically:

```{r}
stats_effect <- res_indo_simple$stats[res_indo_simple$stats$variable == "rx1_indomethacin",]
main_eff_coverage <- empirical_coverage(stats_effect, width = c(0.5,0.9, 0.95))
main_eff_coverage
```

so we would expect e.g. the 95% CI for the main effect to correspond to roughly
`r round(100 * main_eff_coverage$ci_low[3])`% - `r round(100 * main_eff_coverage$ci_high[3])`% credible interval of a fully Bayesian treatment - using more simulations would
let us make this more precise, but for now we are happy that there clearly is no
substantial discrepancy.

We may want to also look at how tight estimates we get - looking directly gives us a
somewhat unhelpful plot:

```{r indo_simple_sim_est_ugly}
plot_sim_estimated(res_indo_simple)
```

There is a simulation where the posterior uncertainty is very large.
This corresponds to observed data where the outcome is the same for all
rows where the treatment was used:

```{r}
biggest_sd_sim <- res_indo_simple$stats$sim_id[
  which.max(res_indo_simple$stats$sd)]
table(datasets_indo_simple$generated[[biggest_sd_sim]][c("outcome", "rx")])
```

Filtering the extreme simulations out,  we see that most commonly, we
get a decently precise estimate.

```{r indo_simple_sim_est}
stats_filtered <- res_indo_simple$stats[res_indo_simple$stats$sd < 200, ]
plot_sim_estimated(stats_filtered, alpha = 0.3)

```

### Badly-informed model

But the simple approximation does not work everytime. Let's say
we want to increase the precision of our main effect estimate
by adjusting for some factors that we believe are associated with the outcome:
the study site, gender and age of the patients. 
Since one site has only 3 patients, we'll remove it from the simulations.
Additionally, we'll standardize the age to be centered at 50 and divide by
10 to make the $N(0,1)$ prior on the age coefficient have a sensible scale.
To make matters worse, we further subsample the data to contain only 100 rows.

```{r}
set.seed(21645222)

indo_rct_complex <- droplevels(
  medicaldata::indo_rct[medicaldata::indo_rct$site != "4_Case",])
rows_to_keep <- sample(1:nrow(indo_rct_complex), size = 100)
indo_rct_complex <- indo_rct_complex[rows_to_keep,]
indo_rct_complex$age <- (indo_rct_complex$age - 50) / 10
indo_rct_complex$risk <- indo_rct_complex$risk - 3
indo_rct_complex$type <- factor(indo_rct_complex$type, ordered = TRUE)

formula_indo_complex <- outcome ~ rx + site + gender + age + risk
datasets_indo_complex <- generate_datasets(SBC_generator_function(
  generator_single_logistic, 
  formula = formula_indo_complex,
  template_data = indo_rct_complex),
  n_sims = 500) 

backend_indo_complex <- SBC_backend_glm(formula = formula_indo_complex, family = "binomial") 
```


Now we are ready to run SBC:

```{r}
res_indo_complex <- compute_SBC(datasets_indo_complex, backend_indo_complex,
                                   cache_mode = "results", 
                                   cache_location = file.path(cache_dir,"indo_complex"))
```

We see the work we've done for diagnostics previously paid off as we are notified 
of some problems.

Additionally the rank plots shows some miscalibration, most pronounced for the 
`site3_UK` coefficients:

```{r indo_complex_ranks}
plot_rank_hist(res_indo_complex)
plot_ecdf_diff(res_indo_complex)
```

What happens is that many of the simulations result in extremely wide uncertainties
around some of the coefficients, making the modest simulated values fall almost exactly
in the middle - this then results in the overabundance of ranks around the midpoint.
A fully Bayesian fit would regularize the uncertainties and avoid this problem.

The main effect of interest (`rx1_indomethacin`) is however still reasonably well calibrated


```{r}
stats_effect <- res_indo_complex$stats[res_indo_complex$stats$variable == "rx1_indomethacin",]
main_eff_coverage <- empirical_coverage(stats_effect, width = c(0.5,0.9, 0.95))
main_eff_coverage
```

### Badly-informed model, narrow priors

The above case aimed at making the normal approximation to the posterior problematic.
We can obviously make things worse by also introducing a strong prior, 
concentrating away from zero which we'll
do here:

```{r}
set.seed(1685554)
datasets_indo_complex_narrow <- generate_datasets(SBC_generator_function(
  generator_single_logistic, 
  formula = formula_indo_complex,
  template_data = indo_rct_complex,
  intercept_prior_loc = 3,
  intercept_prior_width = 0.5,
  predictor_prior_loc = c(-2, 2),
  predictor_prior_width = 0.5),
  n_sims = 500) 
```

```{r}
res_indo_complex_narrow <- compute_SBC(datasets_indo_complex_narrow, backend_indo_complex,
                                   cache_mode = "results", 
                                   cache_location = file.path(cache_dir,"indo_complex_narrow"))
```

This is enough to make basically all the variables poorly calibrated:

```{r indo_complex_narrow_ranks}
plot_rank_hist(res_indo_complex_narrow)
plot_ecdf_diff(res_indo_complex_narrow)

```

### Well-informed model, narrow priors

To make the analysis complete, we'll also return to the simple, well-informed
model, but use narrow priors. 

```{r}
set.seed(3289542)
datasets_indo_simple_narrow <- generate_datasets(SBC_generator_function(
  generator_single_logistic, 
  formula = formula_indo_simple,
  template_data = medicaldata::indo_rct,
  intercept_prior_loc = 3,
  intercept_prior_width = 0.5,
  predictor_prior_loc = c(-2, 2),
  predictor_prior_width = 0.5),
  n_sims = 500) 
```

```{r}
res_indo_simple_narrow <- compute_SBC(datasets_indo_simple_narrow, backend_indo_simple,
                                   cache_mode = "results", 
                                   cache_location = file.path(cache_dir,"indo_simple_narrow"))
```

Turns out that in this case, the likelihood is sometimes not enough to completely overwhelm
the prior and the main treatment effect is poorly calibrated:

```{r indo_simple_narrow_ranks}
plot_rank_hist(res_indo_simple_narrow)
plot_ecdf_diff(res_indo_simple_narrow)

```

## Conclusions

Implementing a minimal support for SBC using an additional algorithm requires writing
three relatively simple functions. If you happen to wrap an algorithm, we'd be
happy to accept a pull request with your backend at https://github.com/hyunjimoon/SBC/.

We've also shown that in some cases, `glm` can actually be a pretty good 
approximation to a fully Bayesian treatment of an equivalent model. Unfortunately,
the approximation does not work always so well - but SBC can tell us when it
works and when it fails.
