---
title: "Discovering indexing errors with SBC"
output: html_notebook
---

```{r setup, include = FALSE}
library(SBC); 
library(ggplot2)

use_cmdstanr <- TRUE # Set to false to use rstan instead

if(use_cmdstanr) {
  library(cmdstanr)
} else {
  library(rstan)
}

```

Below are three different Stan codes for implementing a simple linear regression. Not all of them are correct - can you see which are wrong?

```{r}
stan_code_1 <- "
data {
  int<lower=0> N;   // number of data items
  int<lower=0> K;   // number of predictors
  matrix[N, K] x;   // predictor matrix
  vector[N] y;      // outcome vector
}
parameters {
  real alpha;           // intercept
  vector[K] beta;       // coefficients for predictors
  real<lower=0> sigma;  // error scale
}
model {
  vector[N] mu = rep_vector(alpha, N);
  for(i in 1:K) {
    for(j in 1:N) {
      mu[j] += beta[i] * x[j, i];
    }
  }
  y ~ normal(mu, sigma);  // likelihood
  alpha ~ normal(0, 5);
  beta ~ normal(0, 1);
  sigma ~ normal(0, 2);
}
"

stan_code_2 <- "
data {
  int<lower=0> N;   // number of data items
  int<lower=0> K;   // number of predictors
  matrix[N, K] x;   // predictor matrix
  vector[N] y;      // outcome vector
}
parameters {
  real alpha;           // intercept
  vector[K] beta;       // coefficients for predictors
  real<lower=0> sigma;  // error scale
}
model {
  vector[N] mu;
  for(i in 1:N) {
    mu[i] = alpha;
    for(j in 1:K) {
      mu[i] += beta[j] * x[j, j];
    }
  }
  y ~ normal(mu, sigma);  // likelihood
  alpha ~ normal(0, 5);
  beta ~ normal(0, 1);
  sigma ~ normal(0, 2);
}
"

stan_code_3 <- "
data {
  int<lower=0> N;   // number of data items
  int<lower=0> K;   // number of predictors
  matrix[N, K] x;   // predictor matrix
  vector[N] y;      // outcome vector
}
parameters {
  real alpha;           // intercept
  vector[K] beta;       // coefficients for predictors
  real<lower=0> sigma;  // error scale
}
model {
  y ~ normal(transpose(beta) * transpose(x) + alpha, sigma);  // likelihood
  alpha ~ normal(0, 5);
  beta ~ normal(0, 1);
  sigma ~ normal(0, 2);
}
"




if(use_cmdstanr) {
  stan_model_1 <- cmdstan_model(write_stan_file(stan_code_1))
  stan_model_2 <- cmdstan_model(write_stan_file(stan_code_2))
  stan_model_3 <- cmdstan_model(write_stan_file(stan_code_3))

  backend_1 <- cmdstan_sample_SBC_backend(stan_model_1, iter_warmup = 200, iter_sampling = 200)
  backend_2 <- cmdstan_sample_SBC_backend(stan_model_2, iter_warmup = 200, iter_sampling = 200)
  backend_3 <- cmdstan_sample_SBC_backend(stan_model_3, iter_warmup = 200, iter_sampling = 200)
} else {
  stan_model_1 <- stan_model(model_code = stan_code_1)
  stan_model_2 <- stan_model(model_code = stan_code_2)
  stan_model_3 <- stan_model(model_code = stan_code_3)

  
  backend_1 <- rstan_sample_SBC_backend(stan_model_1, iter = 400, warmup = 200)
  backend_2 <- rstan_sample_SBC_backend(stan_model_2, iter = 400, warmup = 200)
  backend_3 <- rstan_sample_SBC_backend(stan_model_3, iter = 400, warmup = 200)
}

```

If you can, good for you! If not, don't worry, SBC can help you spot coding problems, so let's simulate
data and test all three models against simulated data.

```{r}
set.seed(5666024)

generator_func <- function(N, K) {
  x <- matrix(rnorm(N * K), nrow = N, ncol = K)
  alpha <- rnorm(1, mean = 0, sd = 1)
  beta <- rnorm(K, mean = 0, sd = 1)
  sigma <- abs(rnorm(1, mean = 0, sd = 2))
  
  y <- array(NA_real_, N)
  for(n in 1:N) {
    mu <- alpha
    for(k in 1:K) {
      mu <- mu + x[n,k] * beta[k]
    }
    y[n] <- rnorm(1, mu, sigma) 
  }
  
  list(
    parameters = list(
      alpha = alpha,
      beta = beta,
      sigma = sigma),
    generated = list(
      N = N, K = K,
      x = x, y = y
    )
  )
  
}

datasets <- generate_datasets(function_SBC_generator(generator_func, N = 100, K = 2), 10)

```

```{r}
results_1 <- compute_results(datasets, backend_1)
results_2 <- compute_results(datasets, backend_2)
results_3 <- compute_results(datasets, backend_3)
```

```{r}
for(p in colnames(results_1$ranks)) {
  print(plot_ecdf_diff(results_1, p) + ggtitle(paste0("Model 1: ", p)))
}

```

As far as a quick SBC can see the first code is OK. You could verify further with more iterations
but we tested the model for you and it is OK (although the implementation is not the best one).


```{r}
for(p in colnames(results_2$ranks)) {
  print(plot_ecdf_diff(results_2, p) + ggtitle(paste0("Model 2: ", p)))
}

```

But the second model is actually not looking good.     In fact the problem is the line
`mu[i] += beta[j] * x[j, j];` which should have `x[i, j]` instead. We see that this 
propagates most strongly to the `sigma` parameter (reusing the same `x` element leads to more similar predictions for each row, so `sigma` needs to be inflated to accommodate this)


```{r}
for(p in colnames(results_3$ranks)) {
  print(plot_ecdf_diff(results_3, p) + ggtitle(paste0("Model 3: ", p)))
}


```


And the third model looks OK once again - and in fact we are pretty sure it is also completely correct.
