---
title: Rejection sampling in simulations
author: "Martin Modrák"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Rejection sampling in simulations}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

In some cases, one may want to exclude extreme simulations from SBC (e.g. because
those simulations create divergences). It is best to use
prior predictive checks to examine your priors and change them
to avoid extremes in the simulated data. In some cases, this may however be impractical/impossible to 
do via prior choice - one example are regression coefficients, where
once we have many predictors, any independent prior that is not very strict
will lead to unrealistic predictions. Joint priors are needed in such case, but
those are not well understood and easy to use. See 
[Paul Bürkner's talk on SBC StanConnect](https://www.youtube.com/watch?v=SbgAMkN18dA&t=2340s)
for more context.

An alternative is to use _rejection sampling_ i.e. we repeatedly generate a 
simulation and only accept it when it passes a certain condition we impose
(e.g. that no observed count is larger than $10^8$).
But does rejection sampling when generating simulations affect the validity of SBC?

It turns out that it does not as long as the rejection criterion only
uses observed data and not the unobserved variables. 

We'll first walk through the math and then show examples of both OK and problematic
rejection sampling.


## The math

Let $\mathtt{accept}(y)$ be the probability the the simulated data $y$ is accepted. Note that $\mathtt{accept}$  uses only data as input and would usually be a 0-1 function if you have a clear idea what a "bad" dataset looks like, but could be probabilistic if you're relying on finicky diagnostics. 

We define a variable $a \sim \text{Bernoulli}(\mathtt{accept}(y))$. Given the parameter space $\Theta$ and a specific $\theta \in \Theta$, this implies a joint distribution $\pi(\theta, y, a)$ that factorizes as $\pi(\theta, y, a) = \pi(a|y)\pi(y | \theta)\pi(\theta)$. We can then look at the posterior conditional on accepting a dataset to see the claimed invariance:

$$
\begin{equation}
\pi(\theta | y, a = 1) = \frac{\pi(a = 1 | y) \pi(y | \theta)\pi(\theta)}{\int_\Theta \mathrm{d}\tilde\theta \: \pi(a = 1 | y) \pi(y | \tilde\theta)\pi(\tilde\theta)} = 
\frac{\pi(y | \theta)\pi(\theta)}{\int_\Theta \mathrm{d}\tilde\theta \: \pi(y | \tilde\theta)\pi(\tilde\theta)} = \pi(\theta | y)
\end{equation}
$$


So whether we take rejection into account or not, the model will match the generating process. 
However, if $\mathtt{accept}$ also depended on $\theta$, it would no longer contribute a constant
and we'll get a mismatch between the generator and model.

## Practical examples

So let's see if that also happens in practice. Let's setup our environment:

```{r setup, message=FALSE,warning=FALSE, results="hide"}
library(SBC)

use_cmdstanr <- getOption("SBC.vignettes_cmdstanr", TRUE) # Set to false to use rstan instead

if(use_cmdstanr) {
  library(cmdstanr)
} else {
  library(rstan)
  rstan_options(auto_write = TRUE)
}

library(posterior)

library(future)
plan(multisession) 

options(SBC.min_chunk_size = 10)

# Setup caching of results
if(use_cmdstanr) {
  cache_dir <- "./_rejection_sampling_SBC_cache"
} else {
  cache_dir <- "./_rejection_sampling_rstan_SBC_cache"
}
if(!dir.exists(cache_dir)) {
  dir.create(cache_dir)
}

theme_set(theme_minimal())
```


We'll use a very simple model throughout this vignette:


```{r, comment = ""}
cat(readLines("stan/rejection_sampling.stan"), sep = "\n")
```

```{r}
if(use_cmdstanr) {
  backend <- SBC_backend_cmdstan_sample(cmdstan_model("stan/rejection_sampling.stan"), iter_warmup = 800, iter_sampling = 800)
} else {
  backend <- SBC_backend_rstan_sample(stan_model("stan/rejection_sampling.stan"), iter = 1600, warmup = 800)
}
```

### No rejections

First, we'll use a generator that matches the model exactly.

```{r}
N <- 10
generator <- SBC_generator_function(function() {
   mu <- rnorm(1, 0, 2)
   list(
     variables = list(mu = mu),
     generated = list(N = N, y = rnorm(N, mu, 1))
   )
})
```

So we expect the SBC to pass even with a large number of fits.

```{r}
set.seed(2323455)
datasets <- generate_datasets(generator, 1000)
```

```{r}
results <- compute_SBC(datasets, backend, keep_fits = FALSE, 
                    cache_mode = "results", 
                    cache_location = file.path(cache_dir, "no_rejections"))
```

```{r results1_plots}
plot_ecdf_diff(results)
plot_rank_hist(results)
```

Indeed, all looks good.

### Rejection based on unobserved variables

Now let us modify the generator to reject based on values of an unobserved variable. 

```{r}
generator_reject_unobserved <- SBC_generator_function(function() {
   repeat {
    mu <- rnorm(1, 0, 2)
    if(mu > 3) {
      break
    }
   }
   list(
     variables = list(mu = mu),
     generated = list(N = N, y = rnorm(N, mu, 1))
   )
})
```

We don't even need to run very many fits to see the problem.

```{r}
set.seed(21455)
datasets_reject_unobserved <- generate_datasets(generator_reject_unobserved, 200)
```

```{r}
results_reject_unobserved <- compute_SBC(datasets_reject_unobserved, backend, keep_fits = FALSE, 
                    cache_mode = "results", 
                    cache_location = file.path(cache_dir, "reject_unobserved"))
```

```{r reject_unobserved_plots}
plot_ecdf_diff(results_reject_unobserved)
plot_rank_hist(results_reject_unobserved)
```

Indeed, we see a clear failure.

### Rejecting based on data

But what if we reject based on the values of data? This should in theory result in just
a constant change in posterior density and not affect SBC. (SBC will however then check only the 
non-rejected parts of the data space). We will do a relatively aggressive rejection scheme (reject more than 50% of simulations).

```{r}
generator_reject_y <- SBC_generator_function(function() {
   repeat {
    mu <- rnorm(1, 0, 2)
    y <- rnorm(N, mu, 1)
    if(mean(y) > 5) {
      break
    }
   }
   list(
     variables = list(mu = mu),
     generated = list(N = N, y = y)
   )
})
```

```{r}
set.seed(369654)
datasets_reject_y <- generate_datasets(generator_reject_y, 1000)
```

```{r}
results_reject_y <- compute_SBC(datasets_reject_y, backend, keep_fits = FALSE, 
                    cache_mode = "results", 
                    cache_location = file.path(cache_dir, "reject_y"))
```

```{r reject_y_plot}
plot_rank_hist(results_reject_y)
plot_ecdf_diff(results_reject_y)
```

We see that even with quite heavy rejection based on y, SBC to a high resolution passes.

## Take home message

If our priors can sometimes result in simulated data that is unrealistic, but we are unable to 
specify a better prior directly (e.g. because we would need to define some sort of joint prior),
we can use rejection sampling to prune unrealistic simulations as long as we only filter by the observed
data and don't directly use any unobserved variable values. Notably, filtering based on divergences or other
fitting issues is also just a function of data and thus permissible. The resulting SBC will however provide guarantees
only for data that would not be rejected by the same criteria.
