---
title: "Small model implementation workflow"
author: "Martin Modrák"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Small model implementation workflow}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

Here we describe a complete process to iteratively build and validate the _implementation_ 
of a non-trivial, but still relatively small model. 
This is not a full [Bayesian Workflow](https://arxiv.org/abs/2011.01808), instead
the process described here can be thought of as a subroutine in the full workflow:
here we take a relatively precise description of a model as input and try to produce
a Stan program that implements this model. Once we have a Stan program we trust, 
it is still necessary to validate its fit to actual data and other properties,
which may trigger a need to change the model. At this point you may want to
go back to simulations and make sure the modified model is implemented correctly.

The workflow described here focuses on small models. 
"Small" means that the model is relatively fast to fit and we don't have to worry about computation too much.
Once running ~100 fits of the model becomes too costly, there are additional tricks and considerations
that we hope to delve into in a "Building a complex model" vignette (which currently doesn't exist).
Still many of the approaches here also apply to complex models (especially starting small and building smaller submodels separately), and with proper separation of the model into submodels,
one can validate big chunks of Stan code while working with small models only.

We expect the reader to be familiar with basics of the package. If not,
check out the ["basic_usage" vignette](https://hyunjimoon.github.io/SBC/articles/SBC.html).

## Our goal

The example we'll investigate is building a two-component Poisson mixture,
where the mixing ratio is allowed to vary with some predictors while the means
of the components are the same for all observations.
A somewhat contrived real world situation where this could be a useful model:
there are two sub-species of an animal that are hard to observe directly, but leave
droppings (poop) behind, that we can find. Further, we know the subspecies differ in the 
average number of droppings they leave at one place. So we can take the number of droppings as a noisy
information about which subspecie was present at given location.
We observe the number of droppings at multiple locations and record some environmental covariates about the locations
(e.g. temperature, altitude) and want to learn something about the association between
those covariates and the prevalence of either subspecie.

## Big picture

This model naturally decomposes into two submodels: 

1) the mixture submodel where the mixing ratio is the same
for all observations

2) a beta regression where we take covariates and make a prediction of a probability,
assuming we (noisily) observe the probability.

It is good practice to start small and implement and validate each of those 
submodels separately and then put them together and validate the bigger model.
This makes is substantially easier to locate bugs. 
You'll notice that the process ends up involving a lot of steps, 
but the fact is that we still ignore all the completely invalid models I
created while writing this vignette (typos, compile errors, dimension mismatches, ...). Developing models you
can trust is hard work. More experienced users can definitely make bigger steps at once,
but we strongly discourage anyone from writing a big model in one go. 
My experience is that whenever I try to do this, the model breaks, is impossible
to debug and then I end up breaking it down anyway.

Let's setup and get our hands dirty:

```{r setup, message=FALSE,warning=FALSE, results="hide"}
library(SBC)
library(ggplot2)

use_cmdstanr <- getOption("SBC.vignettes_cmdstanr", TRUE) # Set to false to use rstan instead

if(use_cmdstanr) {
  library(cmdstanr)
} else {
  library(rstan)
  rstan_options(auto_write = TRUE)
}

library(bayesplot)
library(posterior)

library(future)
plan(multisession) 

options(SBC.min_chunk_size = 5)

# Setup caching of results
if(use_cmdstanr) {
  cache_dir <- "./_small_model_worklow_SBC_cache"
} else {
  cache_dir <- "./_small_model_worklow_rstan_SBC_cache"
}
if(!dir.exists(cache_dir)) {
  dir.create(cache_dir)
}

theme_set(theme_minimal())
```

```{r enable progressr, eval = FALSE}
# Run this # Run this _in the console_ to report progress for all computations to report progress for all computations
# see https://progressr.futureverse.org/ for more options
progressr::handlers(global = TRUE)
```


## Mixture submodel

There is a good [guide to mixtures](https://mc-stan.org/docs/2_27/stan-users-guide/mixture-modeling-chapter.html) in the Stan user's guide.
Following the user's guide would save us from a lot of mistakes, but for the sake of example, we will pretend we didn't really read it - and we'll see the problems can be discovered 
via simulations.

So this is our first try at implementing the mixture submodel:


```{r, comment = ""}
cat(readLines("small_model_workflow/mixture_first.stan"), sep = "\n")
```


```{r}
if(use_cmdstanr) {
  model_first <- cmdstan_model("small_model_workflow/mixture_first.stan")
  backend_first <- SBC_backend_cmdstan_sample(model_first) 
} else {
  model_first <- stan_model("small_model_workflow/mixture_first.stan")
  backend_first <- SBC_backend_rstan_sample(model_first) 
}
```

And this is our code to simulate data for this model:

```{r}
generator_func_first <- function(N) {
  mu1 <- rnorm(1, 3, 1)
  mu2 <- rnorm(1, 3, 1)
  theta <- runif(1)
  
  y <- numeric(N)
  for(n in 1:N) {
    if(runif(1) < theta) {
      y[n] <- rpois(1, exp(mu1))
    } else {
      y[n] <- rpois(1, exp(mu2))
    }
  }
  
  list(
    variables = list(
      mu1 = mu1,
      mu2 = mu2,
      theta = theta
    ),
    generated = list(
      N = N,
      y = y
    )
  )
}

generator_first <- SBC_generator_function(generator_func_first, N = 50)
```

Let's start with just a single simulation:

```{r}
set.seed(68455554)
datasets_first <- generate_datasets(generator_first, 1)
```

```{r}
results_first <- compute_SBC(datasets_first, backend_first, 
                    cache_mode = "results", 
                    cache_location = file.path(cache_dir, "mixture_first"))
```

Oh, we have convergence problems, let us examine the pairs plots

```{r mixture_first_convergence}
# Fixing the condition for above/over diagonal chains, in a minority
# of runs the plot shows the problem less clearly, as discussed at
# https://github.com/stan-dev/bayesplot/issues/132
p_cond <- pairs_condition(chains = list(c(1,3), c(2,4)))
if(use_cmdstanr) {
  mcmc_pairs(results_first$fits[[1]]$draws(), condition = p_cond)
} else {
  mcmc_pairs(results_first$fits[[1]], condition = p_cond)  
}
```

One thing that stands out is that either `mu1` is tightly determined and `mu2` is allowed the full prior range or the other way around. We also don't learn anything about theta.

This might be puzzling but relates to bad usage of `log_mix`. The thing is that  `poisson_log_lpmf(y | mu1)` returns a single number - the total log likelihood of all elements of `y` given `mu1`. And thus we are building a mixture where either all observations are from the first component or all are from the second component. To implement mixture where each observation is allowed to come from a different component, we need to loop over observations and do a separate `log_mix` call for each.

More details on the mathematical background are explained in the ["Vectorizing mixtures"](https://mc-stan.org/docs/2_27/stan-users-guide/vectorizing-mixtures.html) section of Stan User's guide.

### Fixing mixture

So we've fixed the `log_mix` problem and this is our new model:

```{r, comment = ""}
cat(readLines("small_model_workflow/mixture_fixed_log_mix.stan"), sep = "\n")
```

```{r}
if(use_cmdstanr) {
  model_fixed_log_mix <- cmdstan_model("small_model_workflow/mixture_fixed_log_mix.stan")
  backend_fixed_log_mix <- SBC_backend_cmdstan_sample(model_fixed_log_mix)
} else {
  model_fixed_log_mix <- stan_model("small_model_workflow/mixture_fixed_log_mix.stan")
  backend_fixed_log_mix <- SBC_backend_rstan_sample(model_fixed_log_mix)
  
}
```

So let's try once again with the same single simulation:

```{r}
results_fixed_log_mix <- compute_SBC(datasets_first, backend_fixed_log_mix, 
                    cache_mode = "results", 
                    cache_location = file.path(cache_dir, "mixture_fixed_log_mix"))
```

No warnings this time. We look at the stats:

```{r}
results_fixed_log_mix$stats
```

We see nothing obviously wrong, the posterior means are relatively close to simulated values (as summarised by the z-scores) - no variable is clearly ridiculously misfit. So let's run a few more iterations.

```{r}
set.seed(8314566)
datasets_first_10 <- generate_datasets(generator_first, 10)
```

```{r}
results_fixed_log_mix_2 <- compute_SBC(datasets_first_10, backend_fixed_log_mix, 
                    cache_mode = "results", 
                    cache_location = file.path(cache_dir, "mixture_fixed_log_mix_2"))
```

So there are some problems - we have quite a bunch of high R-hat and low ESS values. This is the distribution of all rhats:

```{r mixture_fixed_log_mix_rhat}
hist(results_fixed_log_mix_2$stats$rhat)

```

Let's examine a single pairs plot:

```{r mixture_fixed_log_mix_pairs}
if(use_cmdstanr) {
  mcmc_pairs(results_fixed_log_mix_2$fits[[1]]$draws())
} else {
  mcmc_pairs(results_fixed_log_mix_2$fits[[1]])  
}
```

We clearly see two modes in the posterior. And upon reflection, we can see why: swapping `mu1` with `mu2` while also changing `theta` for `1 - theta` gives _exactly_ the same likelihood - because the ordering does not matter. A more detailed explanation of this type of problem is at https://betanalpha.github.io/assets/case_studies/identifying_mixture_models.html

### Fixing ordering

We can easily fix the ordering of the `mu`s by using the `ordered` built-in type.

```{r, comment = ""}
cat(readLines("small_model_workflow/mixture_fixed_ordered.stan"), sep = "\n")
```


```{r}
if(use_cmdstanr) {
  model_fixed_ordered <- cmdstan_model("small_model_workflow/mixture_fixed_ordered.stan")
  backend_fixed_ordered <- SBC_backend_cmdstan_sample(model_fixed_ordered) 
} else {
  model_fixed_ordered <- stan_model("small_model_workflow/mixture_fixed_ordered.stan")
  backend_fixed_ordered <- SBC_backend_rstan_sample(model_fixed_ordered)   
}
```
We also need to update the generator to match the new names and ordering constant:

```{r}
generator_func_ordered <- function(N) {
  # If the priors for all components of an ordered vector are the same
  # then just sorting the result of a generator is enough to create
  # a valid draw from the ordered vector prior
  mu <- sort(rnorm(2, 3, 1)) 
  theta <- runif(1)
  
  y <- numeric(N)
  for(n in 1:N) {
    if(runif(1) < theta) {
      y[n] <- rpois(1, exp(mu[1]))
    } else {
      y[n] <- rpois(1, exp(mu[2]))
    }
  }
  
  list(
    variables = list(
      mu = mu,
      theta = theta
    ),
    generated = list(
      N = N,
      y = y
    )
  )
}

generator_ordered <- SBC_generator_function(generator_func_ordered, N = 50)
```

We are kind of confident (and the model fits quickly), so we'll already start with 10 simulations.

```{r}
set.seed(3785432)
datasets_ordered_10 <- generate_datasets(generator_ordered, 10)
```


```{r}
results_fixed_ordered <- compute_SBC(datasets_ordered_10, backend_fixed_ordered, 
                    cache_mode = "results", 
                    cache_location = file.path(cache_dir, "mixture_fixed_ordered"))
```

Now some fits still produce problematic Rhats or divergent transitions, let's browse the `$backend_diagnostics` (which contain Stan-specific diagnostic values) to see which simulations are causing problems:

```{r}
results_fixed_ordered$backend_diagnostics
```

One of the fits has quite a lot of divergent transitions. Let's look at the pairs plot for the model:

```{r mixture_fixed_ordered_pairs}
problematic_fit_id <- 2
problematic_fit <- results_fixed_ordered$fits[[problematic_fit_id]]
if(use_cmdstanr) {
  mcmc_pairs(problematic_fit$draws(), np = nuts_params(problematic_fit))
} else {
  mcmc_pairs(problematic_fit, np = nuts_params(problematic_fit))
}
```

There is a lot of ugly stuff going on. Notably, one can notice that the posterior of theta is bimodal, preferring either almost 0 or almost 1 - and when that happens, the mean of one of the components is almost unconstrained. 
Why does that happen? The key to the answer is in the simulated values for the component means:

```{r}
subset_draws(datasets_ordered_10$variables, draw = problematic_fit_id)
```

We were unlucky enough to simulate data where both components have almost the same mean and thus we are actually looking at data that is not really a mixture. Mixture models can misbehave badly in such cases (see once again the [case study by Mike Betancourt](https://betanalpha.github.io/assets/case_studies/identifying_mixture_models.html#5_singular_components_and_computational_issues) for a bit more detailed dive into this particular problem).

### Fixing degenerate components?

What to do about this? Fixing the model to handle such cases gracefully is hard. But the problem is basically our prior - we want to express that (since we are fitting a two component model), we don't expect the means to be too similar. So if we can change our simulation to avoid this, we'll be able to proceed with SBC. If such a pattern appeared in real data, we would still have a problem, but we would notice thanks to the diagnostics.

This can definitely be done. But another way is to just ignore the simulations that had divergences for SBC calculations. It turns out that if we remove simulations in a way that only depends on the observed data (and not on unobserved variables), the SBC identity is preserved and we can use SBC without modifications. The resulting check is however telling us something only for data that were not rejected. In this case this is not a big issue: if a fit had divergent transitions, we would not trust it anyway, so removing fits with divergent transitions is not such a big deal.

For more details see the [`rejection_sampling`](https://hyunjimoon.github.io/SBC/articles/rejection_sampling.html) vignette.

So let us subset the results to avoid divergences:

```{r}
sim_ids_to_keep <- 
  results_fixed_ordered$backend_diagnostics$sim_id[
    results_fixed_ordered$backend_diagnostics$n_divergent == 0]

# Equivalent tidy version if you prefer
# sim_ids_to_keep <- results_fixed_ordered$backend_diagnostics %>% 
#   dplyr::filter(n_divergent == 0) %>%
#   dplyr::pull(sim_id)


results_fixed_ordered_subset <- results_fixed_ordered[sim_ids_to_keep]
summary(results_fixed_ordered_subset)
```

This gives us no obvious problems.

```{r mixture_fixed_ordered_subset_results}
plot_rank_hist(results_fixed_ordered_subset)
plot_ecdf(results_fixed_ordered_subset)
```

Since we now have only `r length(results_fixed_ordered_subset)` simulations, it is not surprising that we are still left with a huge uncertainty about the actual coverage of our posterior intervals - we can see that in a plot:

```{r mixture_fixed_ordered_subset_coverage}
plot_coverage(results_fixed_ordered_subset)
```

The coverage plot shows the observed coverage of central posterior intervals of varying width and the associated uncertainty (black + grey), the blue line represents perfect calibration.

Or investigate numerically.

```{r}
coverage <- empirical_coverage(results_fixed_ordered_subset$stats, width = c(0.5,0.9,0.95))
coverage
```

```{r, echo=FALSE}
theta_90_coverage_string <- paste0(round(100 * as.numeric(
  coverage[coverage$variable == "theta" & coverage$width == 0.9, c("ci_low","ci_high")])),
  "%",
  collapse = " - ")
```


We can clearly see that while there are no terrible errors, a quite big miscalibration is still
consistent with the SBC results so far, for example the 90% posterior interval for `theta`
could (as far as we know) contain `r theta_90_coverage_string` of the true values.
That's not very reassuring.


So we can run for more iterations - to reduce memory consumption, we set `keep_fits = FALSE`.
You generally don't want to do this unless you are really short on memory, as it makes you unable to inspect any problems in your fits:

```{r}
set.seed(54987622)
datasets_ordered_100 <- generate_datasets(generator_ordered, 100)
results_fixed_ordered_100 <- compute_SBC(datasets_ordered_100, backend_fixed_ordered, 
                    keep_fits = FALSE, cache_mode = "results", 
                    cache_location = file.path(cache_dir, "mixture_fixed_ordered_100"))
```

Once again we subset to keep only non-divergent fits - this also removes all the problematic Rhats and ESS.

```{r}
sim_ids_to_keep <- 
  results_fixed_ordered_100$backend_diagnostics$sim_id[
    results_fixed_ordered_100$backend_diagnostics$n_divergent == 0]

# Equivalent tidy version
# sim_ids_to_keep <- results_fixed_ordered_100$backend_diagnostics %>% 
#   dplyr::filter(n_divergent == 0) %>%
#   dplyr::pull(sim_id)


results_fixed_ordered_100_subset <- results_fixed_ordered_100[sim_ids_to_keep]
summary(results_fixed_ordered_100_subset)
```


And we can use `bind_results` to combine the new results with the previous fits to not waste our computational effort.

```{r results_fixed_ordered_combined_results}
results_fixed_ordered_combined <- bind_results(
  results_fixed_ordered_subset, results_fixed_ordered_100_subset)

plot_rank_hist(results_fixed_ordered_combined)
plot_ecdf_diff(results_fixed_ordered_combined)
```

Seems fairly well within the expected bounds. We could definitely run more iterations if we wanted to have a more strict check, but for now, we are happy and the remaining uncertainty about the coverage of our posterior intervals is no longer huge, so it is highly unlikely there is some big bug lurking down there. While we see a potential problem where the coverage for `mu[1]` and `mu[2]` is no longer consistent with perfect calibration, the `ecdf_diff` plot takes precedence as the uncertainty in the coverage plot is only approximate and we thus cannot take it too seriously (see `help("empirical_coverage")` for some more details).


```{r results_fixed_ordered_combined_results_coverage}
plot_coverage(results_fixed_ordered_combined)
```


Note: it turns out that extending the model to more components becomes somewhat tricky as the model can become sensitive to initialization. Also the problems with data that can be explained by fewer components than the model assumes become more prevalent.

## Beta regression submodel

Let's move to the beta regression submodel of our model. After spending a bunch of time
implementing this, I realized, that maybe treating this as a logistic regression submodel would have been wiser (and sufficient). But I am gonna keep it in - it just demonstrates that a real workflow can be messy and let's us show some additional classes of problems and how they manifest in SBC.


Checking the wiki page for Beta distribution, we notice that it has two parameters, both bounded to be positive. So our first attempt at beta regression just creates two linear predictors - one for each parameter of the distribution. We then exponentiate the predictors to make them positive and we have a model:

```{r, comment = ""}
cat(readLines("small_model_workflow/beta_first.stan"), sep = "\n")
```

```{r}
if(use_cmdstanr) {
  model_beta_first <- cmdstan_model("small_model_workflow/beta_first.stan")
  backend_beta_first <- SBC_backend_cmdstan_sample(model_beta_first) 
} else {
  model_beta_first <- stan_model("small_model_workflow/beta_first.stan")
  backend_beta_first <- SBC_backend_rstan_sample(model_beta_first)   
}
```

We also write a matching generator (microoptimization tip: I usually write Stan models first so that I can work on the generator code while the Stan model compiles):

```{r}
generator_func_beta_first <- function(N_obs, N_predictors) {
  repeat {
    beta <- matrix(rnorm(N_predictors * 2, 0, 1), nrow = 2, ncol = N_predictors)
  
    x <- matrix(rnorm(N_predictors * N_obs, 0, 1), nrow = N_predictors, ncol = N_obs)
    x[1, ] <- 1 # Intercept
  
    y <- array(NA_real_, N_obs)
      
    for(n in 1:N_obs) {
      linpred <- rep(0, 2)
      for(c in 1:2) {
        for(p in 1:N_predictors) {
          linpred[c] <- linpred[c] + x[p, n] * beta[c, p]
        }
      }
      y[n] <- rbeta(1, exp(linpred[1]), exp(linpred[2]))
    }
    if(all(y < 1 - 1e-12)) {
      break;
    }
  }
    
  list(
    variables = list(
      beta = beta
    ),
    generated = list(
      N_obs = N_obs,
      N_predictors = N_predictors,
      y = y,
      x = x
    )
  )
}

generator_beta_first <- SBC_generator_function(generator_func_beta_first, N_obs = 50, N_predictors = 3)
```

One thing to note is that we add a rejection sampling step - we repeatedly generate simulations,
until we find one without `y` values very close to 1. Those can be problematic as they
can be rounded to 1 when the data for Stan is written to disk. And exact 1 is impossible with the Beta likelihood and the model will fail. Rejecting the simulation due to this criterion is quite rare and in fact, it does not threaten the validity of the SBC procedure (at least to the extent our real data also don't contain such extreme values) - for more details see the [`rejection_sampling`](https://hyunjimoon.github.io/SBC/articles/rejection_sampling.html) vignette.

We'll start with 10 simulations once again.

```{r}
set.seed(3325488)
datasets_beta_first <- generate_datasets(generator_beta_first, 10)
```



```{r}
results_beta_first_10 <- compute_SBC(datasets_beta_first, backend_beta_first, 
                    cache_mode = "results", 
                    cache_location = file.path(cache_dir, "beta_first_10"))
```

We get a single fit with divergent transitions and the ranks look mostly OK:

```{r beta_first_results}
plot_rank_hist(results_beta_first_10)
plot_ecdf(results_beta_first_10)
```

Let's inspect the pairs plot for the offending fit:

```{r beta_first_10_pairs}
if(use_cmdstanr) {
  mcmc_pairs(results_beta_first_10$fits[[3]]$draws())
} else {
  mcmc_pairs(results_beta_first_10$fits[[3]])
}
```

This is a very crowded plot and it is hard to resolve details, but we see some correlations between the corresponding beta elements (e.g. `beta[1,1]` and `beta[2,1]`), let's have a closer look and show the same pairs plot for five of our fits:

```{r beta_first_10_pairs_subset}
for(i in 1:5) {
  fit <- results_beta_first_10$fits[[i]]
  if(use_cmdstanr) {
    pairs_input <- fit$draws()
  } else {
    pairs_input <- fit
  }
  print(mcmc_pairs(pairs_input, pars = c("beta[1,1]", "beta[2,1]","beta[1,2]", "beta[2,2]")))
}
```

Turns out the correlations are in all fits, although sometimes they are relatively weak and the sampler is able to handle the posterior, it is potentially troubling. The main issue is that we plan to integrate this model with other submodels and problems that can be tolerated in a single submodel might interact with other submodels and make the model intractable.


We can even understand the reason for the positive correlation - it is because predicted means of our response beta distribution is the `exp(linpred[1,]) / ( exp(linpred[1,]) + exp(linpred[2,]))` vector.
Increasing both linear predictors by the same amount results in the same predicted means for all elemntes of `x` (but different predicted variances). And changing the two corresponding
`beta` elements has exactly this effect - the two linear predictor  for any `x` increase by the same amount.
In this case, the mean is more constrained than variance of the response, so the individual
predictor values for the first shape parameter are allowed to vary quite a bit as long as their counterpart for the second shape parameter increases as well, keeping the predicted mean the same and showing our uncertainty about the variance. And this is how we get this ridge.


### Parametrizing the beta distribution via mean

The simplest way to resolve the issue with the correlations is to explicitly parametrize the beta distribution by its mean ($0 < \mu < 1$). The more common parametrization then adds a precision parameter ($\phi > 0$), so we then have $y \sim \mathrm{Beta}(\mu \phi, (1 - \mu) \phi)$

This also makes much more sense for the bigger task - combining with the mixture submodel, as we really want to predict just a single probability. So we'll rewrite our predictors to predict only the logit of the mean (as in logistic regression) and keep the precision as a constant between observations. We could definitely also decide whether to keep the full flexibility and allow predictors for precision, we just don't do it here.

This is then our updated model:

```{r, comment = ""}
cat(readLines("small_model_workflow/beta_precision.stan"), sep = "\n")
```

```{r}
if(use_cmdstanr) {
  model_beta_precision <- cmdstan_model("small_model_workflow/beta_precision.stan")
  backend_beta_precision <- SBC_backend_cmdstan_sample(model_beta_precision) 
} else {
  model_beta_precision <- stan_model("small_model_workflow/beta_precision.stan")
  backend_beta_precision <- SBC_backend_rstan_sample(model_beta_precision) 
  
}
```

And we need to update the generator to match:

```{r}
generator_func_beta_precision <- function(N_obs, N_predictors) {
  repeat {
    beta <- rnorm(N_predictors, 0, 1)
    phi <- rlnorm(1, 3, 1)
  
    x <- matrix(rnorm(N_predictors * N_obs, 0, 1), nrow = N_predictors, ncol = N_obs)
    x[1, ] <- 1 # Intercept
  
    y <- array(NA_real_, N_obs)
      
    for(n in 1:N_obs) {
      linpred <- 0
      for(p in 1:N_predictors) {
        linpred <- linpred + x[p, n] * beta[p]
      }
      mu <- plogis(linpred)
      y[n] <- rbeta(1, mu * phi, (1 - mu) * phi)
    }
    if(all(y < 1 - 1e-12)) {
      break;
    }
  }
    
  list(
    variables = list(
      beta = beta,
      phi = phi
    ),
    generated = list(
      N_obs = N_obs,
      N_predictors = N_predictors,
      y = y,
      x = x
    )
  )
}

generator_beta_precision <- 
  SBC_generator_function(generator_func_beta_precision, N_obs = 50, N_predictors = 3)
```


Starting with 10 simulations:

```{r}
set.seed(46988234)
datasets_beta_precision_10 <- generate_datasets(generator_beta_precision, 10)
```



```{r}
results_beta_precision_10 <- compute_SBC(datasets_beta_precision_10, backend_beta_precision, 
                    cache_mode = "results", 
                    cache_location = file.path(cache_dir, "beta_precision_10"))
```

No big problems from the fit and the plots:

```{r results_beta_precision_10_plots}
plot_rank_hist(results_beta_precision_10)
plot_ecdf(results_beta_precision_10)
```

So we'll run 90 more iterations and combine them with the previous results:

```{r}
set.seed(2136468)
datasets_beta_precision_90 <- generate_datasets(generator_beta_precision, 90)
results_beta_precision_90 <- compute_SBC(
  datasets_beta_precision_90, backend_beta_precision,
  keep_fits = FALSE, cache_mode = "results", 
  cache_location = file.path(cache_dir, "beta_precision_90"))


results_beta_precision_100 <-
  bind_results(
    results_beta_precision_10,
    results_beta_precision_90
    )

datasets_beta_precision_100 <- bind_datasets(datasets_beta_precision_10, datasets_beta_precision_90)
```


```{r results_beta_precision_100_plots}
plot_rank_hist(results_beta_precision_100)
plot_ecdf_diff(results_beta_precision_100)
```

The plots don't look terrible, but the `beta[2]` and especially the `phi` variable show slight problems.

So we look back at our model code and note that we forgot to put any prior on `phi`!
Mismatches in priors between the model and the simulator are unfortunately often not very well visible for SBC and can require a lot of simulations to discover (see the [`limits_of_SBC`](https://hyunjimoon.github.io/SBC/articles/limits_of_SBC.html) vignette for more detailed discussion)

### Adding missing prior

So we add the missing prior to the model:

```{r, comment = ""}
cat(readLines("small_model_workflow/beta_precision_fixed_prior.stan"), sep = "\n")
```

```{r}
if(use_cmdstanr) {
  model_beta_precision_fixed_prior <-
    cmdstan_model("small_model_workflow/beta_precision_fixed_prior.stan")
  backend_beta_precision_fixed_prior <- SBC_backend_cmdstan_sample(model_beta_precision_fixed_prior) 
} else {
  model_beta_precision_fixed_prior <-
    stan_model("small_model_workflow/beta_precision_fixed_prior.stan")
  backend_beta_precision_fixed_prior <- SBC_backend_rstan_sample(model_beta_precision_fixed_prior)   
}
```

And recompute for all 100 simulations at once (as we don't expect adding prior to introduce huge problems).

```{r}
results_beta_precision_fixed_prior <- 
  compute_SBC(datasets_beta_precision_100, backend_beta_precision_fixed_prior, 
                    keep_fits = FALSE, cache_mode = "results", 
                    cache_location = file.path(cache_dir, "beta_precision_fixed_prior"))
```

```{r results_beta_precision_fixed_prior_plots}
plot_rank_hist(results_beta_precision_fixed_prior)
plot_ecdf_diff(results_beta_precision_fixed_prior)
```

Diagnostic plots are looking good! So we add 100 more simulations:


```{r}
set.seed(1233845)
datasets_beta_precision_100b <- generate_datasets(generator_beta_precision, 100)
results_beta_precision_fixed_prior_200 <-
  bind_results(
    results_beta_precision_fixed_prior,
    compute_SBC(datasets_beta_precision_100b, backend_beta_precision_fixed_prior, 
                    keep_fits = FALSE, cache_mode = "results", 
                    cache_location = file.path(cache_dir, "beta_precision_fixed_prior_2")))
```

```{r results_beta_precision_fixed_prior_200_plots}
plot_rank_hist(results_beta_precision_fixed_prior_200)
plot_ecdf_diff(results_beta_precision_fixed_prior_200)
```

Yeah, still looking good. And we can see that the empirical coverage of our central intervals
is in quite tight agreement with theory:

```{r results_beta_precision_fixed_prior_200_coverage}
plot_coverage(results_beta_precision_fixed_prior_200)
```

So for now we are also happy about the beta regression submodel.

## Putting it together

We are finally ready to make a first attempt at the full model:

```{r, comment = ""}
cat(readLines("small_model_workflow/combined_first.stan"), sep = "\n")
```

```{r}
if(use_cmdstanr) {
  model_combined <- cmdstan_model("small_model_workflow/combined_first.stan")
  backend_combined <- SBC_backend_cmdstan_sample(model_combined)
} else {
  model_combined <- stan_model("small_model_workflow/combined_first.stan")
  backend_combined <- SBC_backend_rstan_sample(model_combined)  
}
```

And this is our generator for the full model:

```{r}
generator_func_combined <- function(N_obs, N_predictors) {
  # If the priors for all components of an ordered vector are the same
  # then just sorting the result of a generator is enough to create
  # a valid draw from the ordered vector prior
  mu <- sort(rnorm(2, 3, 1)) 
  
  beta <- rnorm(N_predictors, 0, 1)

  x <- matrix(rnorm(N_predictors * N_obs, 0, 1), nrow = N_predictors, ncol = N_obs)
  x[1, ] <- 1 # Intercept

  y <- array(NA_real_, N_obs)

  for(n in 1:N_obs) {
    linpred <- 0
    for(p in 1:N_predictors) {
      linpred <- linpred + x[p, n] * beta[p]
    }
    theta <- plogis(linpred)
    
    if(runif(1) < theta) {
      y[n] <- rpois(1, exp(mu[1]))
    } else {
      y[n] <- rpois(1, exp(mu[2]))
    }
    
  }


  list(
    variables = list(
      beta = beta,
      mu = mu
    ),
    generated = list(
      N_obs = N_obs,
      N_predictors = N_predictors,
      y = y,
      x = x
    )
  )
}

generator_combined <- SBC_generator_function(generator_func_combined, N_obs = 50, N_predictors = 3)
```

We are confident (and the fits are fast anyway), so we start with 200 simulations:

```{r}
set.seed(5749955)
dataset_combined <- generate_datasets(generator_combined, 200)
```

```{r}
results_combined <- compute_SBC(dataset_combined, backend_combined, 
                    keep_fits = FALSE, cache_mode = "results", 
                    cache_location = file.path(cache_dir, "combined"))
```
We get some amount of divergent transitions, but the ranks look pretty good:

```{r results_combined_plots}
plot_rank_hist(results_combined)
plot_ecdf_diff(results_combined)
```

Indeed it seems the model works pretty well.

### Adding rejection sampling

As done previously, we could just exclude the fits that had divergences, but just to complete our tour of possibilities, we'll show one more option to dealing with this type of problem.

The general idea is that although we might not want to/be able to express our prior belief about the model (here that the two mixture components are distinct) by priors on model parameters, we still may be able to express our prior belief about the data itself.

And it turns out that if we remove simulations that don't meet a certain condition imposed on the observed data, the implied prior on parameters becomes an additive constant and we can use exactly the same model to fit only the non-rejected simulations. Note that this does not hold if we rejected simulations based on some unobserved variables - for more details see the [`rejection_sampling`](https://hyunjimoon.github.io/SBC/articles/rejection_sampling.html) vignette.

The main advantage is that if we can do this, we can avoid wasting computation on fitting data that would likely produce divergences anyway. The downside is that it means we no longer have a guarantee the model works for non-rejected data, so we need to check if the data we want to analyze would not be rejected by our criterion.

How to build such a criterion here? We'll note that for Poisson-distributed variables the ratio of mean to variance (a.k.a the Fano factor) is always 1. So if the components are too similar, the data should resemble a Poisson distribution and have Fano factor of 1, while if the components are distinct the Fano factor will be larger. 

Below is a plot of fano factors versus the number of divergences we've seen:

```{r combined_fanos}
fanos <- vapply(dataset_combined$generated, 
                function(dataset) { var(dataset$y) / mean(dataset$y) }, 
                FUN.VALUE = 0)
plot(fanos, results_combined$backend_diagnostics$n_divergent)
```

All the divergence are for low fano factors - this is the histogram of Fano factor for diverging fits:

```{r fanos_divergent}
hist(fanos[results_combined$backend_diagnostics$n_divergent > 0])
fano_threshold <- 1.6
```

So what we'll do is that we'll reject any simulation where the observed data have Fano factor < `r fano_threshold`. In practice a simple way to implement this is to wrap our generator code in a loop and break from the loop only when the generated data meet our criteria (i.e. is not rejected). This is our code:

```{r}
generator_func_combined_reject <- function(N_obs, N_predictors) {
  if(N_obs < 5) {
    stop("Too low N_obs for this simulator")
  }
  repeat {
    # If the priors for all components of an ordered vector are the same
    # then just sorting the result of a generator is enough to create
    # a valid draw from the ordered vector prior
    mu <- sort(rnorm(2, 3, 1)) 
    
    beta <- rnorm(N_predictors, 0, 1)

    x <- matrix(rnorm(N_predictors * N_obs, 0, 1), nrow = N_predictors, ncol = N_obs)
    x[1, ] <- 1 # Intercept
  
    y <- array(NA_real_, N_obs)

    for(n in 1:N_obs) {
      linpred <- 0
      for(p in 1:N_predictors) {
        linpred <- linpred + x[p, n] * beta[p]
      }
      theta <- plogis(linpred)
      
      if(runif(1) < theta) {
        y[n] <- rpois(1, exp(mu[1]))
      } else {
        y[n] <- rpois(1, exp(mu[2]))
      }
      
    }
    if(var(y) / mean(y) > fano_threshold) {
      break;
    }
  }
    
  list(
    variables = list(
      beta = beta,
      mu = mu
    ),
    generated = list(
      N_obs = N_obs,
      N_predictors = N_predictors,
      y = y,
      x = x
    )
  )
}

generator_combined_reject <- 
  SBC_generator_function(generator_func_combined_reject, N_obs = 50, N_predictors = 3)
```

We'll once again fit our model to 200 simulations:

```{r}
set.seed(44685226)
dataset_combined_reject <- generate_datasets(generator_combined_reject, 200)
```



```{r}
results_combined_reject <- compute_SBC(dataset_combined_reject, backend_combined, 
                    keep_fits = FALSE, cache_mode = "results", 
                    cache_location = file.path(cache_dir, "combined_reject"))
```

No more divergences! And the ranks look nice.

```{r combined_reject_plots}
plot_rank_hist(results_combined_reject)
plot_ecdf_diff(results_combined_reject)
```

And our coverage is pretty tight:

```{r combined_reject_coverage}
plot_coverage(results_combined_reject)
```

Below we show the uncertainty for two variables and some widths of central posterior intervals numerically:

```{r results_combined_reject_coverage}
stats_subset <- results_combined_reject$stats[
  results_combined_reject$stats$variable %in% c("beta[1]", "mu[1]"),]

empirical_coverage(stats_subset, c(0.25,0.5,0.9,0.95))
```

Maybe we think the remaining uncertainty is too big, so we'll run 300 more simulations, just to be sure:

```{r}
set.seed(1395367854)
dataset_combined_reject_more <- generate_datasets(generator_combined_reject, 300) 
results_combined_reject_more <- bind_results(
  results_combined_reject,
  compute_SBC(dataset_combined_reject_more, backend_combined, 
                    keep_fits = FALSE, cache_mode = "results", 
                    cache_location = file.path(cache_dir, "combined_reject_more"))
)
```
We get some very small number of problematic fits, which we can ignore in this volume (but probably more aggresive rejection sampling would remove those as well).

Our plots and coverage are now pretty decent:

```{r results_combined_reject_more_plots}
plot_rank_hist(results_combined_reject_more)
plot_ecdf_diff(results_combined_reject_more)
```

```{r results_combined_reject_more_coverage}
plot_coverage(results_combined_reject_more)
```

```{r}
stats_subset <- results_combined_reject_more$stats[
  results_combined_reject_more$stats$variable %in% c("beta[1]", "mu[2]"),]
empirical_coverage(stats_subset, c(0.25,0.5,0.9,0.95))
```

This actually shows a limitation of the coverage results - for `mu[2]` the approximate CI for coverage excludes exact calibration for a bunch of intervals, but above we see that the more trustworthy `plot_ecdf_diff` is not showing a problem (although there is some tendency towards slight underdispersion).

Still, this might warrant further investigation if small discrepancies in `mu` are considered important, if we are interested only in the `beta` coefficients, we can stay assured that their calibration is pretty good. We give you our word that we ran additional simulations and the discrepancy disappears.

Finally, we can also use this simulation exercise to understand what would we be likely to learn from an experiment matching the simulations (50 observations, 3 predictors) and plot the true values (simulated by the generator) against estimated mean + 90% posterior credible interval:

```{r sim_estimated_final}
plot_sim_estimated(results_combined_reject_more, alpha = 0.2)
```

We see that we get very precise information about `mu` and a decent picture about all `beta` elements, but the reamining uncertainty is large. We could for example compute the probability
that the posterior 90% interval for `beta[1]` excludes zero, i.e. that we learn something about the sign of `beta[1]`: 

```{r}
stats_beta1 <- 
  results_combined_reject_more$stats[
    results_combined_reject_more$stats$variable == "beta[1]",]

mean(sign(stats_beta1$q5) == sign(stats_beta1$q95))
```
Turns out the probability is only around 50%. Depending on your aims, this might be a reason to plan for a larger sample size!

## Take home message

There are couple lessons I hope this exercise showed: First, building models 
you can trust is hard work and it is very easy to make mistakes. Despite the
models presented here being relatively simple, diagnosing the problems in them
was not straightforward and required non-trivial background knowledge. For this
reason, moving in small steps during model development is crucial and can save
you time as diagnosing the same problems in a 300-line Stan model with 50 parameters 
can be  basically impossible.

We also hope we convinced you that the SBC package lets you get high-quality 
information from your simulation efforts and not only diagnose problems but also 
get some sort of assurance in the end that your model is at least pretty
close to your simulator.

And that's it for this vignette, thanks for staying until the end and hope the workflow ideas will be useful for you!



